### 2024-09-25 paper/paper.qmd
Update paper.qmd for the data part

---
title: "My title"
subtitle: "My subtitle if needed"
author: 
  - Jiwon Choi
thanks: "Code and data are available at: https://github.com/jwonc4602/Analysis-of-Motor-Vehicle-Collision-Causes."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

#### Workspace set-up ####

## Load packages
library(tidyverse)
library(knitr)
library(here)
library(dplyr)
library(ggplot2)
library(readr)
library(gridExtra)
library(grid)

#read in cleaned data #
#total_cleaned_data
cleaned_data <- read.csv(here::here("data/analysis_data/total_cleaned_data.csv"))
show_col_types = FALSE

# human factors data
human_factors_data <- read.csv(here::here("data/analysis_data/human_factors_data.csv"))
show_col_types = FALSE

# environmental factors data
environmental_factors_data <- read.csv(here::here("data/analysis_data/environmental_factors_data.csv"))
show_col_types = FALSE

# result data
result_data <- read.csv(here::here("data/analysis_data/result_data.csv"))
show_col_types = FALSE
```

# Introduction

You can and should cross-reference sections and sub-sections. We use @citeR and @rohan.

The remainder of this paper is structured as follows. @sec-data....

# Data {#sec-data}

## Data Source

This research is based on a dataset published by Toronto Police Services, which is accessible through the City of Toronto’s OpenDataToronto Library [@rOpenDataToronto]. The specific dataset includes all traffic collisions events in Toronto where a person was either Killed or Seriously Injured (KSI) since 2006.

## Data Collection

The methodology for gathering data is structured to ensure a detailed and current perspective of shelter and overnight service. The data was compiled and examined using the R statistical programming software [@r], supplemented by various tools such as `tidyverse` [@rTidyverse], `ggplot2` [@rGgplot2], `dplyr` [@rDplyr], `readr` [@rReadr], `tibble` [@rTibble], `janitor` [@rJanitor], `knitr` [@rknitr], and `here` [@rHere].

This traffic accident data consists of 18,957 entries from a citywide accident survey, capturing detailed information about various factors involved in traffic incidents. The raw dataset includes 50 variables, ranging from geographic data, such as the street where the accident occurred, to behavioral data, such as driver condition and alcohol involvement. This dataset records accidents over multiple years and provides insights into environmental conditions, road types, and the actions of drivers and pedestrians at the time of the incidents. The goal of this dataset is to provide a comprehensive view of traffic accidents for research and policy analysis related to urban traffic safety and accident prevention.

The cleaned data is divided into three datasets, and the first one focuses on human factors related to traffic collisions, specifically the behaviors and conditions of drivers that may have contributed to accidents. The four main categories—'speeding', 'aggressive_and_disturbed_driving', 'alcohol', and 'disability'—were selected because these are well-known, critical factors that significantly increase the risk of traffic accidents. Each factor is evaluated as either "None" or "Yes", indicating the absence or presence of that factor in the collision. These categories were chosen because they represent preventable human actions or conditions that are highly correlated with the likelihood and severity of traffic incidents. (see @tbl-cleaned-human-factors-data)

1. **speeding**: Refers to collisions where speeding was a contributing factor. Speeding reduces a driver's ability to react quickly to road conditions or obstacles, and it increases the severity of crashes when they occur, making it one of the primary causes of traffic collisions.

2. **aggressive_and_disturbed_driving**: This category examines whether aggressive behavior, such as reckless driving, or distracted driving (like texting or talking on the phone) was involved in the collision. Aggressive and distracted driving are known to impair judgment and reaction time, leading to accidents that could otherwise have been avoided.

3. **alcohol**: Evaluates if alcohol was a factor in the collision. Alcohol impairs motor skills and decision-making, and is a well-documented contributor to severe and fatal traffic accidents. This category is crucial for understanding the role of substance use in driver behavior.

4. **disability**: Assesses if any medical or physical disability played a role in the collision. Disabilities, whether temporary or permanent, can affect a driver's ability to control the vehicle or respond to changing road conditions, making this an important factor to consider in the context of traffic safety.

These categories were selected because they represent key human factors that have a direct impact on traffic safety, and addressing them can help reduce both the frequency and severity of accidents.

```{r}
#| message: false
#| echo: false
#| label: fig-frequency-factors-tall
#| fig-cap: Frequency of human factors with thinner numbers and taller graph

# Convert relevant columns to factors and reorder them for better readability
human_factors_data <- human_factors_data %>%
  mutate(speeding = factor(speeding, levels = c("None", "Yes")),
         aggressive_and_disturbed_driving = factor(aggressive_and_disturbed_driving, levels = c("None", "Yes")),
         alcohol = factor(alcohol, levels = c("None", "Yes")),
         disability = factor(disability, levels = c("None", "Yes")))

# Frequency summary for each factor
factor_summary <- human_factors_data %>%
  summarise(
    speeding = sum(speeding == "Yes"),
    aggressive_and_disturbed_driving = sum(aggressive_and_disturbed_driving == "Yes"),
    alcohol = sum(alcohol == "Yes"),
    disability = sum(disability == "Yes")
  ) %>%
  tidyr::gather(key = "factor", value = "count")

# Split the category name for 'aggressive_and_disturbed_driving'
factor_summary <- factor_summary %>%
  mutate(factor = str_replace_all(factor, "aggressive_and_disturbed_driving", "aggressive\nand disturbed driving"))

# Reorder factors based on count for better visualization
factor_summary <- factor_summary %>%
  mutate(factor = forcats::fct_reorder(factor, count, .desc = TRUE))

# Bar plot with thinner numbers and taller graph
ggplot(factor_summary, aes(x = factor, y = count, fill = factor)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = count), vjust = -0.5, nudge_y = 200, size = 3, fontface = "plain", color = "black") +  # Thinner numbers
  theme_minimal() +
  labs(title = "Frequency of Human Factors in Traffic Collisions",
       x = "Human Factors",
       y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Adjust x-axis labels
  coord_cartesian(ylim = c(0, 11000))  # Adjust the y-axis limit to make the graph taller

```

```{r}
#| message: false
#| echo: false
#| label: fig-trend-over-time
#| fig-cap: Trends of human factors in traffic collisions over time

# Convert the date column to date format
human_factors_data <- human_factors_data %>%
  mutate(date = as.Date(date, format = "%Y-%m-%d"))

# Summarise the presence of each factor by year
trend_data <- human_factors_data %>%
  mutate(year = lubridate::year(date)) %>%
  group_by(year) %>%
  summarise(
    speeding = sum(speeding == "Yes"),
    aggressive_and_disturbed_driving = sum(aggressive_and_disturbed_driving == "Yes"),
    alcohol = sum(alcohol == "Yes"),
    disability = sum(disability == "Yes")
  ) %>%
  tidyr::gather(key = "factor", value = "count", -year)

# Line plot to show trends over time
ggplot(trend_data, aes(x = year, y = count, color = factor)) +
  geom_line(linewidth = 0.8) +
  theme_minimal() +
  labs(title = "Trends in Human Factors Over Time",
       x = "Year",
       y = "Count of Collisions",
       color = "Human Factor")
```

The second dataset focuses on the environmental factors of traffic collisions, capturing the conditions that surround and influence accidents. These factors play a critical role in how drivers respond to the road and other vehicles, and they can either increase or mitigate the risk of collisions. The selected categories—'traffic_control_type', 'visibility', 'light', and 'road_surface_condition'—were chosen because they reflect the external conditions that can directly affect driver behavior and vehicle performance during a collision. Each category contributes to understanding the environment in which the collision occurred and highlights preventable factors or areas for safety improvement. (see @tbl-cleaned-environmental-factors-data)

1. **traffic_control_type**: This category identifies the type of traffic control present at the scene, such as "Stop Sign," "Traffic Signal," "Pedestrian Crossover," or "No Control." These labels were chosen because the presence and type of traffic control significantly influence driver decisions. Proper traffic control can prevent accidents, while its absence or failure can lead to dangerous situations.

2. **visibility**: This category describes the environmental visibility conditions at the time of the collision, such as "Clear," "Fog, Mist, Smoke, Dust," "Rain," or "Drifting Snow." Visibility is critical for a driver’s ability to react to road hazards or changes. Poor visibility, whether caused by weather or other environmental factors, can lead to misjudgments or delays in reaction time, making it an important factor to analyze in traffic collisions.

3. **Light**: This refers to the lighting conditions at the time of the collision, with possibilities such as "Daylight," "Dusk," "Dark," or artificial lighting conditions like "Dark, artificial" or "Dusk, artificial." Lighting impacts how well a driver can see the road, pedestrians, and other vehicles. Dim or changing light conditions can impair a driver’s perception, increasing the likelihood of an accident, which is why this category is crucial for safety assessments.

4. **road_surface_condition**: This captures the state of the road surface, including conditions like "Dry," "Wet," "Loose Snow," "Ice," or "Slush." The condition of the road surface affects the vehicle’s traction and stability, making it one of the key determinants of how well a driver can maintain control of their vehicle. Poor road surface conditions, such as ice or wet roads, are a known cause of accidents and were chosen for their direct impact on vehicle performance.

These categories were selected because they represent the most critical environmental factors that influence traffic safety. Understanding and addressing these factors can improve road designs, traffic control systems, and driver awareness, ultimately reducing collision rates.

```{r}
#| message: false
#| echo: false
#| label: tbl-cleaned-environmental-factors-data
#| tbl-cap: Sample of Environmental Factors of Collision Data

head (environmental_factors_data, 5) |>
  kable(
    col.names = c("date", "traffic_control_type", "visibility", "light", "road_surface_condition"),
  booktabs = TRUE
  )
```

```{r}
#| message: false
#| echo: false
#| label: fig-four-graphs-layout
#| fig-cap: Combined layout with four simplified collision graphs

# Modify traffic_control_type to group categories into 'No Control', 'Stop Sign', 'Traffic Signal', and 'Others'
environmental_factors_data <- environmental_factors_data %>%
  mutate(traffic_control_type_grouped = case_when(
    traffic_control_type == "No Control" ~ "No Control",
    traffic_control_type == "Stop Sign" ~ "Stop Sign",
    traffic_control_type == "Traffic Signal" ~ "Traffic Signal",
    TRUE ~ "Others"
  ))

# Modify visibility to group categories into 'Clear', 'Rain', 'Snow', and 'Others'
environmental_factors_data <- environmental_factors_data %>%
  mutate(visibility_grouped = case_when(
    visibility == "Clear" ~ "Clear",
    visibility == "Rain" ~ "Rain",
    visibility == "Snow" ~ "Snow",
    TRUE ~ "Others"
  ))

# Common theme
common_theme <- theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),  # Slightly larger font for clarity
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    plot.title = element_text(size = 10),
    plot.margin = margin(t = 15, r = 15, b = 15, l = 15)  # Margins for more space
  )

# First plot: Traffic Control Type and Visibility Conditions

# Plot 1: Traffic Control Type
plot1 <- ggplot(environmental_factors_data, aes(x = traffic_control_type_grouped)) +
  geom_bar(fill = "steelblue") +
  geom_text(stat = 'count', aes(label = after_stat(count)), vjust = -0.5, size = 3) +
  common_theme +
  labs(title = "Collisions by Traffic Control Type",
       x = "Traffic Control Type",
       y = "Number of Collisions") +
  coord_cartesian(ylim = c(0, max(table(environmental_factors_data$traffic_control_type_grouped)) * 1.2))

# Plot 2: Visibility Conditions
plot2 <- ggplot(environmental_factors_data, aes(x = visibility_grouped)) +
  geom_bar(fill = "steelblue") +
  geom_text(stat = 'count', aes(label = after_stat(count)), vjust = -0.5, size = 3) +
  common_theme +
  labs(title = "Collisions by Visibility Conditions",
       x = "Visibility",
       y = "Number of Collisions") +
  coord_cartesian(ylim = c(0, max(table(environmental_factors_data$visibility_grouped)) * 1.2))

# Arrange these two plots side by side
first_plot_pair <- grid.arrange(plot1, plot2, ncol = 2, 
                                top = textGrob("Environmental Factors Affecting Collisions", gp = gpar(fontsize = 15, fontface = "bold")))

```

```{r}
#| message: false
#| echo: false
#| label: tbl-cleaned-environmental-factors-data-2
#| tbl-cap: Sample of Environmental Factors of Collision Data 2

# Common theme
common_theme <- theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    axis.title.x = element_text(size = 10),
    axis.title.y = element_text(size = 10),
    plot.title = element_text(size = 10),  # No bold here
    plot.margin = margin(t = 15, r = 15, b = 15, l = 15)  # Margins for more space
  )

# Modify light to group categories into 'Dark', 'Dawn', 'Daylight', 'Dusk', and 'Others'
environmental_factors_data <- environmental_factors_data %>%
  mutate(lighting_grouped = case_when(
    light %in% c("Dark", "Dark, artificial") ~ "Dark",
    light %in% c("Dawn", "Dawn, artificial") ~ "Dawn",
    light %in% c("Daylight", "Daylight, artificial") ~ "Daylight",
    light %in% c("Dusk", "Dusk, artificial") ~ "Dusk",
    light %in% c("None", "Other") ~ "Others"
  ))

# Modify road_surface_condition to group categories into 'Dry', 'Wet', and 'Other'
environmental_factors_data <- environmental_factors_data %>%
  mutate(road_surface_grouped = case_when(
    road_surface_condition == "Dry" ~ "Dry",
    road_surface_condition == "Wet" ~ "Wet",
    TRUE ~ "Other"
  ))

# Second plot: Lighting Conditions and Road Surface Condition

# Plot 3: Lighting Conditions
plot3 <- ggplot(environmental_factors_data, aes(x = lighting_grouped)) +
  geom_bar(fill = "steelblue") +
  geom_text(stat = 'count', aes(label = after_stat(count)), vjust = -0.5, size = 3) +
  common_theme +  # Apply the common theme
  labs(title = "Collisions by Lighting Conditions",
       x = "Lighting Conditions",
       y = "Collisions") +
  coord_cartesian(ylim = c(0, max(table(environmental_factors_data$lighting_grouped)) * 1.2))

# Plot 4: Road Surface Conditions
plot4 <- ggplot(environmental_factors_data, aes(x = road_surface_grouped)) +
  geom_bar(fill = "steelblue") +
  geom_text(stat = 'count', aes(label = after_stat(count)), vjust = -0.5, size = 3) +
  common_theme +  # Apply the common theme
  labs(title = "Collisions by Road Surface Condition",
       x = "Road Surface Condition",
       y = "Collisions") +
  coord_cartesian(ylim = c(0, max(table(environmental_factors_data$road_surface_grouped)) * 1.2))

# Arrange these two plots side by side
second_plot_pair <- grid.arrange(plot3, plot4, ncol = 2, 
                                 top = textGrob("Environmental Factors Affecting Collisions: Part 2", gp = gpar(fontsize = 15, fontface = "bold")))

```

The last dataset focuses on the result data, specifically analyzing the outcomes of traffic collisions in terms of the 'classification_of_accident' and the 'injury'. These categories were selected because they offer essential insights into the real-world consequences of traffic accidents, both in terms of human harm and material damage. By understanding the outcomes, safety measures and interventions can be better designed to prevent severe accidents in the future. (see @tbl-cleaned-result-data)

1. **classification_of_accident**: This category outlines the type of accident that occurred, with classifications such as "Fatal," "Non-Fatal Injury," "None," and "Property Damage Only." These labels were selected because they provide a clear and structured way to assess the severity of the collision. This classification helps to categorize accidents based on their impact, whether they resulted in fatalities, injuries, or just property damage. By understanding the distribution of these classifications, authorities can prioritize safety interventions to reduce fatal or injury-related accidents.

2. **injury**: This category refers to the level of injury sustained in the accident, with possible values including "Fatal," "Major," "Minor," "Minimal," and "None." Injury severity is a key indicator of the human toll of traffic collisions. These specific levels were selected because they cover the full spectrum of injury outcomes, from minor incidents to fatal accidents. This allows for a detailed understanding of the physical impact on those involved in the accidents, facilitating better emergency response planning and public health interventions.

These categories were chosen because they provide crucial information on the aftermath of traffic collisions, helping to identify patterns in accident severity and injury. Understanding these outcomes is vital for designing targeted measures to reduce the occurrence and impact of serious traffic accidents.

```{r}
#| message: false
#| echo: false
#| label: tbl-cleaned-result-data
#| tbl-cap: Sample of Result of Collision Data

head (result_data, 5) |>
  kable(
    col.names = c("date", "classification_of_accident", "injury"),
  booktabs = TRUE
  )
```

# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

# References

### 2024-09-25 scripts/03-test_data.R
Update 03-test_data.R, delete unused variable

@@ -51,34 +51,28 @@ test_results$test_5 <- all(unique(finalized_accident_data$classification_of_acci
 expected_injury <- c("Fatal", "Major", "Minimal", "Minor", "None")
 test_results$test_6 <- all(unique(finalized_accident_data$injury) %in% expected_injury)
 
-# Test 7: 'driver_condition' categories
-expected_driver_conditions <- c("Ability Impaired, Alcohol", "Ability Impaired, Alcohol Over .08", 
-                                "Ability Impaired, Drugs", "Fatigue", "Had Been Drinking", "Inattentive", 
-                                "Medical or Physical Disability", "None", "Normal", "Other", "Unknown")
-test_results$test_7 <- all(unique(finalized_accident_data$driver_condition) %in% expected_driver_conditions)
-
-# Test 8: 'speeding' categories
+# Test 7: 'speeding' categories
 expected_speeding <- c("None", "Yes")
-test_results$test_8 <- all(unique(finalized_accident_data$speeding) %in% expected_speeding)
+test_results$test_7 <- all(unique(finalized_accident_data$speeding) %in% expected_speeding)
 
-# Test 9: 'aggressive_and_disturbed_driving' categories
+# Test 8: 'aggressive_and_disturbed_driving' categories
 expected_aggressive_driving <- c("None", "Yes")
-test_results$test_9 <- all(unique(finalized_accident_data$aggressive_and_disturbed_driving) %in% expected_aggressive_driving)
+test_results$test_8 <- all(unique(finalized_accident_data$aggressive_and_disturbed_driving) %in% expected_aggressive_driving)
 
-# Test 10: 'alcohol' categories
+# Test 9: 'alcohol' categories
 expected_alcohol <- c("None", "Yes")
-test_results$test_10 <- all(unique(finalized_accident_data$alcohol) %in% expected_alcohol)
+test_results$test_9 <- all(unique(finalized_accident_data$alcohol) %in% expected_alcohol)
 
-# Test 11: 'disability' categories
+# Test 10: 'disability' categories
 expected_disability <- c("None", "Yes")
-test_results$test_11 <- all(unique(finalized_accident_data$disability) %in% expected_disability)
+test_results$test_10 <- all(unique(finalized_accident_data$disability) %in% expected_disability)
 
-# Test 12: 'date' format
+# Test 11: 'date' format
 # Check if the 'date' column is in the correct format 'YYYY-MM-DD'
[…]

### 2024-09-26 scripts/02-data_cleaning.R
update on some variables and delete 'injury'

@@ -20,42 +20,62 @@ raw_data <- read_csv("data/raw_data/raw_data.csv")
 cleaned_data <-
   raw_data |>
   janitor::clean_names() |>
-  select(date, traffctl, visibility, light, rdsfcond, acclass, injury, speeding, ag_driv, alcohol, disability) |>
+  select(date, traffctl, visibility, light, rdsfcond, acclass, speeding, ag_driv, alcohol, disability) |>
   tidyr::drop_na()
 
-# Rename multiple columns to understand more easily by the name
-cleaned_data <- cleaned_data |>
-  rename(
-    traffic_control_type = traffctl,
-    road_surface_condition = rdsfcond,
-    classification_of_accident = acclass,
-    aggressive_and_disturbed_driving = ag_driv
-  )
+cleaned_data <- cleaned_data %>%
+  mutate(traffctl = case_when(
+    traffctl == "No Control" ~ "No Control",
+    traffctl == "Traffic Signal" ~ "Traffic Signal",
+    TRUE ~ "Others"
+  ))
+
+# Modify visibility to group categories into 'Clear' or 'Disturbed'
+cleaned_data <- cleaned_data %>%
+  mutate(visibility = case_when(
+    visibility == "Clear" ~ "Clear",
+    TRUE ~ "Disturbed"
+  ))
+
+# Modify light to group categories into 'Dark','Daylight', and 'Dim Light'
+cleaned_data <- cleaned_data %>%
+  mutate(light = case_when(
+    light %in% c("Dark", "Dark, artificial") ~ "Dark",
+    light %in% c("Daylight", "Daylight, artificial") ~ "Daylight",
+    TRUE ~ "Dim Light"
+  ))
+
+# Modify rdsfcond to group categories into 'Dry', 'Wet', and 'Others'
+cleaned_data <- cleaned_data %>%
+  mutate(rdsfcond = case_when(
+    rdsfcond == "Dry" ~ "Dry",
+    rdsfcond == "Wet" ~ "Wet",
+    TRUE ~ "Others"
+  ))
+
+cleaned_data <- cleaned_data %>%
+  mutate(acclass = case_when(
+    acclass == "Fatal" ~ "Fatal",
+    acclass == "Non-Fatal Injury" ~ "Non-Fatal Injury",
+    TRUE ~ "None or Others"
+  ))
 
 # human factors
 human_factors_data <-
   cleaned_data |>
   janitor::clean_names() |>
-  select(date, speeding, aggressive_and_disturbed_driving, alcohol, disability) |>
[…]

### 2024-09-26 scripts/03-test_data.R

@@ -10,7 +10,7 @@
 library(tidyverse)
 
 # Read the finalized data
-finalized_accident_data <- read_csv("data/analysis_data/analysis_data.csv")
+finalized_accident_data <- read_csv("data/analysis_data/total_cleaned_data.csv")
 
 #### Data Cleaning ####
 # Remove leading/trailing spaces in relevant columns
@@ -22,57 +22,49 @@ test_results <- list()
 
 #### Test unique values for categorical columns ####
 
-# Test 1: 'traffic_control_type' categories
-expected_traffic_control_types <- c("No Control", "None", "Pedestrian Crossover", "Police Control",
-                                    "School Guard", "Stop Sign", "Streetcar (Stop for)", "Traffic Controller", 
-                                    "Traffic Gate", "Traffic Signal", "Yield Sign")
-test_results$test_1 <- all(unique(finalized_accident_data$traffic_control_type) %in% expected_traffic_control_types)
+# Test 1: 'traffctl' categories
+expected_traffic_control_types <- c("No Control", "Traffic Signal", "Others")
+test_results$test_1 <- all(unique(finalized_accident_data$traffctl) %in% expected_traffic_control_types)
 
 # Test 2: 'visibility' categories
-expected_visibility <- c("Clear", "Drifting Snow", "Fog, Mist, Smoke, Dust", "Freezing Rain", 
-                         "None", "Other", "Rain", "Snow", "Strong wind")
+expected_visibility <- c("Clear", "Disturbed")
 test_results$test_2 <- all(unique(finalized_accident_data$visibility) %in% expected_visibility)
 
 # Test 3: 'light' categories
-expected_light <- c("Dark", "Dark, artificial", "Dawn", "Dawn, artificial", "Daylight", 
-                    "Daylight, artificial", "Dusk", "Dusk, artificial", "None", "Other")
+expected_light <- c("Dark", "Dim Light", "Daylight")
 test_results$test_3 <- all(unique(finalized_accident_data$light) %in% expected_light)
 
-# Test 4: 'road_surface_condition' categories
-expected_road_surface_conditions <- c("Dry", "Ice", "Loose Sand or Gravel", "Loose Snow", "None", "Other", 
-                                      "Packed Snow", "Slush", "Spilled liquid", "Wet")
-test_results$test_4 <- all(unique(finalized_accident_data$road_surface_condition) %in% expected_road_surface_conditions)
[…]

### 2024-09-26 paper/paper.qmd
Update paper.qmd

@@
 ---
-title: "My title"
-subtitle: "My subtitle if needed"
+title: "Analysis of Motor Vehicle Collision Causes in Toronto from 2006 to 2023"
+subtitle: "Exploring the Interplay of Human Behavior and Environmental Factors in Predicting Accident Severity and High-Risk Scenarios"
 author: 
@@
 date-format: long
-abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
+abstract: "This paper analyzes motor vehicle collisions in Toronto from 2006 to 2023, focusing on the interplay between human behavior and environmental factors.
+The study examines how behaviors such as speeding, aggressive driving, alcohol consumption, and disability contribute to accidents alongside conditions like visibility, lighting, and road surface.
+Findings reveal that while environmental factors can exacerbate risks, human actions, particularly aggressive driving and alcohol use, are the most significant contributors to severe collisions.
+These insights are critical for developing predictive models and preventive strategies that can enhance urban road safety by addressing both driver behavior and infrastructure improvements."
 format: pdf
@@
 
-# result data
-result_data <- read.csv(here::here("data/analysis_data/result_data.csv"))
-show_col_types = FALSE
 ```
@@
 
-You can and should cross-reference sections and sub-sections. We use @citeR and @rohan.
+Motor vehicle collisions remain one of the leading causes of injury and fatality in urban environments, prompting ongoing research into how these incidents can be mitigated.
+In a rapidly growing city like Toronto, understanding the causes of traffic accidents is essential for improving road safety and informing policy.
+This paper investigates traffic collisions in Toronto over a 17-year span, analyzing the combined effects of human behavior and environmental conditions to determine which factors most significantly contribute to the frequency and severity of accidents.
 
-The remainder of this paper is structured as follows. @sec-data....
+Despite extensive research on individual factors like speeding or poor road conditions, there remains a gap in understanding how human behaviors interact with environmental conditions to create high-risk scenarios for accidents.
+Existing studies tend to focus on either human or environmental factors in isolation, missing the crucial interplay between these elements.
+This research aims to fill that gap by examining both human and environmental factors in tandem, providing a more holistic view of traffic accident causes in an urban context.
+
+Using data from Toronto Police Services, this study explores the role of speeding, aggressive driving, alcohol consumption, and disability alongside environmental factors such as visibility, lighting, traffic control, and road surface conditions.
+The findings reveal that aggressive driving and alcohol use, especially under favorable road conditions, are leading contributors to collisions.
+The study also identifies high-risk combinations, such as aggressive driving on dry roads under clear visibility, which result in the highest number of accidents.
+These results are important because they offer a nuanced understanding of traffic safety and suggest that targeted interventions, such as stricter penalties for reckless driving or improved traffic control in key areas, could significantly reduce accidents.
+
+The remainder of this paper is structured as follows.
+Section 2 provides a detailed description of the data and methodology.
+Section 3 presents the analysis of human and environmental factors contributing to traffic collisions.
+Section 4 discusses the results, highlighting high-risk scenarios and potential interventions.
+Finally, Section 5 addresses the limitations of the current study and outlines directions for future research, particularly the need to account for the growing presence of autonomous driving technologies in collision analysis.
 
@@
 
-The cleaned data is divided into three datasets, and the first one focuses on human factors related to traffic collisions, specifically the behaviors and conditions of drivers that may have contributed to accidents.
-The four main categories—'speeding', 'aggressive_and_disturbed_driving', 'alcohol', and 'disability'—were selected because these are well-known, critical factors that significantly increase the risk of traffic accidents.
+The cleaned data is divided into two datasets, and the first one focuses on human factors related to traffic collisions, specifically the behaviors and conditions of drivers that may have contributed to accidents.
+The four main categories—'speeding', 'ag_driv', 'alcohol', and 'disability'—were selected because these are well-known, critical factors that significantly increase the risk of traffic accidents.
 Each factor is evaluated as either "None" or "Yes", indicating the absence or presence of that factor in the collision.
@@
 2.
-**aggressive_and_disturbed_driving**: This category examines whether aggressive behavior, such as reckless driving, or distracted driving (like texting or talking on the phone) was involved in the collision.
+**ag_driv**: This category examines whether aggressive behavior, such as reckless driving, or distracted driving (like texting or talking on the phone) was involved in the collision.
 Aggressive and distracted driving are known to impair judgment and reaction time, leading to accidents that could otherwise have been avoided.
@@
   mutate(speeding = factor(speeding, levels = c("None", "Yes")),
-         aggressive_and_disturbed_driving = factor(aggressive_and_disturbed_driving, levels = c("None", "Yes")),
+         ag_driv = factor(ag_driv, levels = c("None", "Yes")),
          alcohol = factor(alcohol, levels = c("None", "Yes")),
@@
     speeding = sum(speeding == "Yes"),
-    aggressive_and_disturbed_driving = sum(aggressive_and_disturbed_driving == "Yes"),
+    ag_driv = sum(ag_driv == "Yes"),
     alcohol = sum(alcohol == "Yes"),
@@
 
-# Split the category name for 'aggressive_and_disturbed_driving'
+# Split the category name for 'ag_driv'
 factor_summary <- factor_summary %>%
-  mutate(factor = str_replace_all(factor, "aggressive_and_disturbed_driving", "aggressive\nand disturbed driving"))
+  mutate(factor = str_replace_all(factor, "ag_driv", "aggressive\nand disturbed driving"))
 
@@
 
-The second dataset focuses on the environmental factors of traffic collisions, capturing the conditions that surround and influence accidents.
-These factors play a critical role in how drivers respond to the road and other vehicles, and they can either increase or mitigate the risk of collisions.
-The selected categories—'traffic_control_type', 'visibility', 'light', and 'road_surface_condition'—were chosen because they reflect the external conditions that can directly affect driver behavior and vehicle performance during a collision.
-Each category contributes to understanding the environment in which the collision occurred and highlights preventable factors or areas for safety improvement.
+The second dataset focuses on the environmental factors influencing traffic collisions, capturing the conditions that shape driver behavior and contribute to accidents.
+These factors, including traffic control, visibility, lighting, and road surface conditions, play a vital role in either increasing or reducing the risk of collisions.
+The following modifications were made to simplify and group these categories, making the data more consistent and easier to analyze.
+The merged categories provide clearer insights into how these factors affect accidents and highlight areas for safety improvements.
 (see @tbl-cleaned-environmental-factors-data)
@@
 1.
-**traffic_control_type**: This category identifies the type of traffic control present at the scene, such as "Stop Sign," "Traffic Signal," "Pedestrian Crossover," or "No Control." These labels were chosen because the presence and type of traffic control significantly influence driver decisions.
-Proper traffic control can prevent accidents, while its absence or failure can lead to dangerous situations.
+**traffic_control_type**: This category was simplified to include "No Control," "Traffic Signal," and "Others." The "Others" category consolidates less frequent types of traffic control, such as "Stop Sign," "Pedestrian Crossover," and "School Guard." This grouping reflects the most significant distinctions in how traffic is regulated, emphasizing scenarios where the presence or absence of formal traffic control could impact driver behavior.
+By focusing on the broad types of control, this category helps assess how different levels of traffic management influence accident risks.
 
 2.
-**visibility**: This category describes the environmental visibility conditions at the time of the collision, such as "Clear," "Fog, Mist, Smoke, Dust," "Rain," or "Drifting Snow." Visibility is critical for a driver’s ability to react to road hazards or changes.
-Poor visibility, whether caused by weather or other environmental factors, can lead to misjudgments or delays in reaction time, making it an important factor to analyze in traffic collisions.
+**visibility**: The visibility conditions were grouped into "Clear" and "Disturbed." The "Disturbed" category includes various adverse weather conditions such as "Fog, Mist, Smoke, Dust," "Rain," "Snow," and "Strong Wind." This grouping simplifies the analysis by highlighting the key distinction between ideal and hazardous visibility conditions, which directly affect a driver’s ability to see and react to road conditions.
+Poor visibility is a known contributor to accidents, making this distinction critical for identifying environmental hazards.
 
 3.
-**Light**: This refers to the lighting conditions at the time of the collision, with possibilities such as "Daylight," "Dusk," "Dark," or artificial lighting conditions like "Dark, artificial" or "Dusk, artificial." Lighting impacts how well a driver can see the road, pedestrians, and other vehicles.
-Dim or changing light conditions can impair a driver’s perception, increasing the likelihood of an accident, which is why this category is crucial for safety assessments.
+**light**: The lighting conditions were merged into three categories: "Dark" (which includes "Dark" and "Dark, artificial"), "Daylight" (including "Daylight" and "Daylight, artificial"), and "Dim Light" (covering "Dawn," "Dusk," and similar intermediate lighting conditions).
+These groups reflect the significant lighting scenarios that impact driver visibility.
+Dark and dim lighting can reduce reaction times and perception, making these categories essential for understanding how lighting conditions contribute to accidents.
 
 4.
-**road_surface_condition**: This captures the state of the road surface, including conditions like "Dry," "Wet," "Loose Snow," "Ice," or "Slush." The condition of the road surface affects the vehicle’s traction and stability, making it one of the key determinants of how well a driver can maintain control of their vehicle.
-Poor road surface conditions, such as ice or wet roads, are a known cause of accidents and were chosen for their direct impact on vehicle performance.
+**road_surface_condition**: Road surface conditions were consolidated into "Dry," "Wet," and "Others." The "Others" category includes conditions like "Ice," "Loose Snow," and "Slush," which can severely reduce traction.
+Grouping less common conditions into "Others" helps streamline the analysis by focusing on the primary road conditions that affect vehicle control.
+Wet or slippery surfaces are known to increase accident risks, making this distinction important for understanding how road conditions influence collisions.
 
-These categories were selected because they represent the most critical environmental factors that influence traffic safety.
-Understanding and addressing these factors can improve road designs, traffic control systems, and driver awareness, ultimately reducing collision rates.
+These categories were selected and modified to prioritize the most critical environmental factors that affect traffic safety.
+By simplifying and grouping related variables, the data becomes more interpretable, allowing for a clearer focus on how environmental factors like traffic control, visibility, lighting, and road conditions contribute to motor vehicle collisions.
+This approach not only aids in identifying preventable causes but also supports efforts to improve road safety and reduce collision rates through targeted interventions.
 
@@
   kable(
-    col.names = c("date", "traffic_control_type", "visibility", "light", "road_surface_condition"),
+    col.names = c("date", "traffctl", "visibility", "light", "rdsfcond"),
   booktabs = TRUE
@@
 #| fig-cap: Combined layout with four simplified collision graphs
-
-# Modify traffic_control_type to group categories into 'No Control', 'Stop Sign', 'Traffic Signal', and 'Others'
-environmental_factors_data <- environmental_factors_data %>%
-  mutate(traffic_control_type_grouped = case_when(
-    traffic_control_type == "No Control" ~ "No Control",
-    traffic_control_type == "Stop Sign" ~ "Stop Sign",
-    traffic_control_type == "Traffic Signal" ~ "Traffic Signal",
-    TRUE ~ "Others"
-  ))
-
-# Modify visibility to group categories into 'Clear', 'Rain', 'Snow', and 'Others'
-environmental_factors_data <- environmental_factors_data %>%
-  mutate(visibility_grouped = case_when(
-    visibility == "Clear" ~ "Clear",
-    visibility == "Rain" ~ "Rain",
-    visibility == "Snow" ~ "Snow",
-    TRUE ~ "Others"
-  ))
 
@@
 # Plot 1: Traffic Control Type
-plot1 <- ggplot(environmental_factors_data, aes(x = traffic_control_type_grouped)) +
+plot1 <- ggplot(environmental_factors_data, aes(x = traffctl)) +
   geom_bar(fill = "steelblue") +
@@
        y = "Number of Collisions") +
-  coord_cartesian(ylim = c(0, max(table(environmental_factors_data$traffic_control_type_grouped)) * 1.2))
+  coord_cartesian(ylim = c(0, max(table(environmental_factors_data$traffctl)) * 1.2))
 
 # Plot 2: Visibility Conditions
-plot2 <- ggplot(environmental_factors_data, aes(x = visibility_grouped)) +
+plot2 <- ggplot(environmental_factors_data, aes(x = visibility)) +
   geom_bar(fill = "steelblue") +
@@
        y = "Number of Collisions") +
-  coord_cartesian(ylim = c(0, max(table(environmental_factors_data$visibility_grouped)) * 1.2))
+  coord_cartesian(ylim = c(0, max(table(environmental_factors_data$visibility)) * 1.2))
 
@@
 
-# Modify light to group categories into 'Dark', 'Dawn', 'Daylight', 'Dusk', and 'Others'
-environmental_factors_data <- environmental_factors_data %>%
-  mutate(lighting_grouped = case_when(
-    light %in% c("Dark", "Dark, artificial") ~ "Dark",
-    light %in% c("Dawn", "Dawn, artificial") ~ "Dawn",
-    light %in% c("Daylight", "Daylight, artificial") ~ "Daylight",
-    light %in% c("Dusk", "Dusk, artificial") ~ "Dusk",
-    light %in% c("None", "Other") ~ "Others"
-  ))
-
-# Modify road_surface_condition to group categories into 'Dry', 'Wet', and 'Other'
-environmental_factors_data <- environmental_factors_data %>%
-  mutate(road_surface_grouped = case_when(
-    road_surface_condition == "Dry" ~ "Dry",
-    road_surface_condition == "Wet" ~ "Wet",
-    TRUE ~ "Other"
-  ))
-
 # Second plot: Lighting Conditions and Road Surface Condition
@@
 # Plot 3: Lighting Conditions
-plot3 <- ggplot(environmental_factors_data, aes(x = lighting_grouped)) +
+plot3 <- ggplot(environmental_factors_data, aes(x = light)) +
   geom_bar(fill = "steelblue") +
@@
        y = "Collisions") +
-  coord_cartesian(ylim = c(0, max(table(environmental_factors_data$lighting_grouped)) * 1.2))
+  coord_cartesian(ylim = c(0, max(table(environmental_factors_data$light)) * 1.2))
 
 # Plot 4: Road Surface Conditions
-plot4 <- ggplot(environmental_factors_data, aes(x = road_surface_grouped)) +
+plot4 <- ggplot(environmental_factors_data, aes(x = rdsfcond)) +
   geom_bar(fill = "steelblue") +
@@
        y = "Collisions") +
-  coord_cartesian(ylim = c(0, max(table(environmental_factors_data$road_surface_grouped)) * 1.2))
+  coord_cartesian(ylim = c(0, max(table(environmental_factors_data$rdsfcond)) * 1.2))
 
@@
 
-The last dataset focuses on the result data, specifically analyzing the outcomes of traffic collisions in terms of the 'classification_of_accident' and the 'injury'.
-These categories were selected because they offer essential insights into the real-world consequences of traffic accidents, both in terms of human harm and material damage.
-By understanding the outcomes, safety measures and interventions can be better designed to prevent severe accidents in the future.
-(see @tbl-cleaned-result-data)
-
-1.
-**classification_of_accident**: This category outlines the type of accident that occurred, with classifications such as "Fatal," "Non-Fatal Injury," "None," and "Property Damage Only." These labels were selected because they provide a clear and structured way to assess the severity of the collision.
-This classification helps to categorize accidents based on their impact, whether they resulted in fatalities, injuries, or just property damage.
-By understanding the distribution of these classifications, authorities can prioritize safety interventions to reduce fatal or injury-related accidents.
-
-2.
-**injury**: This category refers to the level of injury sustained in the accident, with possible values including "Fatal," "Major," "Minor," "Minimal," and "None." Injury severity is a key indicator of the human toll of traffic collisions.
-These specific levels were selected because they cover the full spectrum of injury outcomes, from minor incidents to fatal accidents.
-This allows for a detailed understanding of the physical impact on those involved in the accidents, facilitating better emergency response planning and public health interventions.
-
-These categories were chosen because they provide crucial information on the aftermath of traffic collisions, helping to identify patterns in accident severity and injury.
-Understanding these outcomes is vital for designing targeted measures to reduce the occurrence and impact of serious traffic accidents.
+Lastly, the **acclass** category classifies the type and severity of traffic collisions with labels such as "Fatal," "Non-Fatal Injury," and "None or Others." The original classifications of "None" and "Property Damage Only" were merged into the "None or Others" category to streamline the data, as both involve accidents without human injury.
+This consolidation was taken to simplify the analysis by focusing on the human impact of accidents.
+"Fatal" represents collisions resulting in severe injury, and "Non-Fatal Injury" indicates incidents causing injury but no fatalities.
+By merging and refining these categories, the focus shifts toward understanding the human consequences of collisions, enabling a more targeted approach to reducing severe injuries and fatalities.
 
@@
 #| echo: false
-#| label: tbl-cleaned-result-data
-#| tbl-cap: Sample of Result of Collision Data
+#| label: fig-acclass-distribution
+#| fig-cap: Distribution of Accident Classifications (acclass)
 
-head (result_data, 5) |>
-  kable(
-    col.names = c("date", "classification_of_accident", "injury"),
-  booktabs = TRUE
-  )
+# Ensure the 'acclass' column is treated as a factor
+cleaned_data <- cleaned_data %>%
+  mutate(acclass = factor(acclass, levels = c("Fatal", "Non-Fatal Injury", "None or Others")))
+
+# Summarise the counts for each 'acclass' category
+acclass_summary <- cleaned_data %>%
+  count(acclass)
+
+# Create the bar plot
+ggplot(acclass_summary, aes(x = acclass, y = n, fill = acclass)) +
+  geom_bar(stat = "identity") +
+  geom_text(aes(label = n), vjust = -0.5, nudge_y = 200, size = 3, fontface = "plain", color = "black") +
+  theme_minimal() +
+  labs(title = "Distribution of Accident Classifications",
+       x = "Accident Classification",
+       y = "Count") +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
+  coord_cartesian(ylim = c(0, max(acclass_summary$n) * 1.1))  # Adjust y-axis limits for better readability
+
 ```
@@
 
-## First discussion point {#sec-first-point}
+## High-Risk Scenarios for Road Accidents
 
-If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.
+Upon further review of the heatmap, the top three combinations of environmental and human factors are evident, and I will revise the analysis accordingly.
 
-## Second discussion point
+### High-Risk Scenarios for Road Accidents
 
-## Third discussion point
+The heatmap highlights the intricate relationship between environmental conditions—such as traffic control, visibility, lighting, and road surface—and human factors like speeding, aggressive driving, alcohol use, and disability.
+This visualization reveals clear trends where specific combinations of these factors lead to the highest rates of road accidents.
+
+The highest-risk combination occurs when **traffic signals are clear, daylight is present, the road is dry, and drivers exhibit aggressive driving behavior**.
+This combination results in the largest number of accidents.
+Even though environmental conditions are favorable with clear daylight and dry roads, aggressive driving significantly increases the accident rate.
+This indicates that aggressive behaviors, such as tailgating or reckless overtaking, are extremely dangerous even in otherwise safe driving conditions.
+To mitigate these risks, more targeted enforcement of speed limits and penalties for aggressive driving may be necessary in areas where accidents are high, despite good road conditions.
+
+A second high-risk combination involves **no traffic control, clear conditions, daylight, and dry roads, combined with aggressive driving and alcohol use**.
+This combination results in high accident counts, emphasizing how dangerous aggressive driving becomes when compounded by alcohol consumption, even in seemingly safe environmental conditions.
+The absence of traffic control in these situations suggests that drivers may feel a false sense of security, which encourages reckless behavior.
+Enhancing traffic control measures, such as adding stop signs, traffic lights, or roundabouts in high-accident zones, could reduce incidents in these cases.
+
+The third-highest combination includes **no traffic control, disturbed visibility, daylight, and wet road conditions** with **no involvement of impaired behaviors**.
+This scenario shows how environmental factors like wet roads and poor visibility increase the risk of accidents, even when human behavior is not a contributing factor.
+The lack of traffic control adds further risk.
+Addressing this issue through better traffic control measures and improving road surface conditions, such as better drainage systems or warning signs in areas prone to heavy rainfall, can help reduce accidents.
+
+The data reveals that while environmental factors such as wet roads, disturbed visibility, and lack of traffic control are significant contributors to accidents, aggressive driving and alcohol use are key human behaviors that exacerbate risks, even in favorable environmental conditions.
+Focusing on infrastructure improvements, such as enhancing traffic control in key areas and stricter enforcement against aggressive driving and alcohol-related offenses, could significantly reduce accident rates in the most hazardous scenarios.
+
+```{r fig.width=12, fig.height=15}
+#| message: false
+#| echo: false
+#| label: fig-hh
+#| fig-cap: Sample of Result of Collision Data
+# Prepare the data by summarizing relevant columns and creating a count of combinations
+heatmap_data <- cleaned_data %>%
+  group_by(traffctl, visibility, light, rdsfcond, speeding, ag_driv, alcohol, disability) %>%
+  summarise(count = n()) %>%
+  ungroup()
+
+# Combine human factor columns into one for the x-axis
+heatmap_data$human_factors <- interaction(heatmap_data$speeding,
+                                          heatmap_data$ag_driv,
+                                          heatmap_data$alcohol,
+                                          heatmap_data$disability)
+
+# Combine environmental factor columns into one for the y-axis
+heatmap_data$environmental_factors <- interaction(heatmap_data$traffctl,
+                                                  heatmap_data$visibility,
+                                                  heatmap_data$light,
+                                                  heatmap_data$rdsfcond)
+
+# Create the heatmap plot with better alignment
+ggplot(heatmap_data, aes(x = human_factors, 
+                         y = environmental_factors, 
+                         fill = count)) +
+  geom_tile(color = "white") +
+  scale_fill_gradient(low = "lightblue", high = "red") +
+  labs(x = "Human Factors (speeding, ag_driv, alcohol, disability)", 
+       y = "Environmental Factors (traffctl, visibility, light, rdsfcond)", 
+       fill = "Count") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 15), # Increase text size and rotate
+        axis.text.y = element_text(size = 17), # Increase text size
+        axis.title = element_text(size = 20), # Increase axis title size
+        plot.title = element_text(size = 22),  # Increase title size
+        legend.title = element_text(size = 20),
+        legend.text = element_text(size = 18),
+        panel.grid.major = element_blank(), # Remove gridlines
+        panel.grid.minor = element_blank()) +
+  ggtitle("Heatmap of Environmental and Human Factors")
+
+```
+
+## Can Environmental and Human Factors Serve as Early Warning Indicators?
+With the advancement of data-driven technologies, it is possible to predict the severity of motor vehicle accidents based on a combination of human behaviors and environmental conditions.
+This discussion focuses on how predictive models can be developed to identify the likelihood of severe accidents using factors like speeding, alcohol consumption, poor road conditions, and low visibility.
+
+For example, we could create a system that collects real-time environmental data (e.g., weather, road surface conditions) and cross-references it with information on traffic behaviors (e.g., average speeding rates) to predict whether an area is at high risk for serious accidents.
+This kind of early warning system could be integrated into navigation apps or smart cars to alert drivers of heightened risks and potentially prevent accidents before they occur.
+
+Predictive analytics can highlight the most dangerous combinations of factors, such as speeding in rainy weather or aggressive driving under poor visibility.
+In addition, machine learning algorithms can continuously improve these predictions by incorporating accident history data, making the model more robust over time.
+This offers the potential to shift from reactive to proactive accident prevention, reducing the burden on healthcare systems and increasing road safety.
 
@@
 
-Weaknesses and next steps should also be included.
+One notable limitation of this analysis is that the dataset only contains information up until 2023.
+While this provides a substantial historical perspective on traffic accidents in Toronto, it may not fully capture recent changes in road usage patterns, traffic enforcement, or driver behaviors that have emerged post-2023.
+For example, the rise of new driving technologies, ride-sharing services, and changes in traffic regulations could have a significant impact on accident rates, but these factors are not reflected in the current dataset.
+As such, the findings based on this data may not completely align with more recent trends, limiting the relevance of the analysis for current or future policy decisions.
+
+While the dataset used in this analysis ends in 2023, a key area for future exploration involves the broader adoption of autonomous or auto-drive technologies, which are not yet accounted for in the data but could significantly impact traffic accident dynamics.
+As these technologies become more prevalent, understanding their effect on traffic safety, accident rates, and accident severity will be critical.
+The absence of this data limits our ability to predict how autonomous driving might influence future accident patterns.
+However, from the figure "fig-trend-over-time," we observe that collisions involving aggressive driving (ag_driv) have decreased over time, which may be indicative of increased use of autonomous or driver-assist technologies that mitigate reckless behavior.
+Future research should incorporate data on autonomous vehicles and their interaction with human-driven cars to better assess how these advancements influence safety and accident risks.
+This could lead to the development of predictive models that combine human and automated driving behaviors, aiding policymakers and urban planners in shaping infrastructure and safety regulations for a future where both types of vehicles coexist.
 

### 2024-09-27 paper/paper.qmd
major update on paper

@@
 date-format: long
-abstract: "This paper analyzes motor vehicle collisions in Toronto from 2006 to 2023, focusing on the interplay between human behavior and environmental factors.
-The study examines how behaviors such as speeding, aggressive driving, alcohol consumption, and disability contribute to accidents alongside conditions like visibility, lighting, and road surface.
-Findings reveal that while environmental factors can exacerbate risks, human actions, particularly aggressive driving and alcohol use, are the most significant contributors to severe collisions.
-These insights are critical for developing predictive models and preventive strategies that can enhance urban road safety by addressing both driver behavior and infrastructure improvements."
+abstract: "This study analyzes motor vehicle collisions in Toronto from 2006 to 2023, focusing on the interaction between human behaviors—such as speeding, aggressive driving, and alcohol use—and environmental conditions like visibility and road surfaces.
+The findings show that while environmental factors increase risks, aggressive driving and alcohol use are the strongest predictors of severe accidents.
+These results highlight the need for targeted interventions in both driver behavior and infrastructure to improve urban traffic safety."
 format: pdf
@@
 
-Motor vehicle collisions remain one of the leading causes of injury and fatality in urban environments, prompting ongoing research into how these incidents can be mitigated.
-In a rapidly growing city like Toronto, understanding the causes of traffic accidents is essential for improving road safety and informing policy.
-This paper investigates traffic collisions in Toronto over a 17-year span, analyzing the combined effects of human behavior and environmental conditions to determine which factors most significantly contribute to the frequency and severity of accidents.
+Motor vehicle collisions are a leading cause of injury and fatality in urban areas, driving ongoing efforts to improve road safety.
+In a rapidly expanding city like Toronto, understanding the causes of traffic accidents is critical for shaping effective policies.
+This paper examines traffic collisions in Toronto over a 17-year period, focusing on how human behavior and environmental conditions together influence the frequency and severity of accidents.
 
-Despite extensive research on individual factors like speeding or poor road conditions, there remains a gap in understanding how human behaviors interact with environmental conditions to create high-risk scenarios for accidents.
-Existing studies tend to focus on either human or environmental factors in isolation, missing the crucial interplay between these elements.
-This research aims to fill that gap by examining both human and environmental factors in tandem, providing a more holistic view of traffic accident causes in an urban context.
+While much research has been done on factors such as speeding or poor road conditions, there is still a lack of understanding regarding the interaction between human behaviors and environmental conditions in creating high-risk situations.
+Most studies focus on these factors separately, missing how they combine to increase accident risks.
+This research addresses that gap by analyzing both human and environmental factors simultaneously, offering a more complete view of what drives traffic accidents in an urban setting.
 
-Using data from Toronto Police Services, this study explores the role of speeding, aggressive driving, alcohol consumption, and disability alongside environmental factors such as visibility, lighting, traffic control, and road surface conditions.
-The findings reveal that aggressive driving and alcohol use, especially under favorable road conditions, are leading contributors to collisions.
-The study also identifies high-risk combinations, such as aggressive driving on dry roads under clear visibility, which result in the highest number of accidents.
-These results are important because they offer a nuanced understanding of traffic safety and suggest that targeted interventions, such as stricter penalties for reckless driving or improved traffic control in key areas, could significantly reduce accidents.
+Using data from Toronto Police Services, this study analyzes the role of human factors like speeding, aggressive driving, alcohol use, and disability alongside environmental factors such as visibility, lighting, traffic control, and road conditions.
+The results show that aggressive driving and alcohol use, even under good road conditions, are major contributors to accidents.
+High-risk scenarios, like aggressive driving on dry roads with clear visibility, emerge as leading causes of collisions.
+These findings suggest targeted actions, such as stricter enforcement against reckless driving and improved traffic controls in key areas, could help reduce accident rates.
 
-The remainder of this paper is structured as follows.
-Section 2 provides a detailed description of the data and methodology.
-Section 3 presents the analysis of human and environmental factors contributing to traffic collisions.
-Section 4 discusses the results, highlighting high-risk scenarios and potential interventions.
-Finally, Section 5 addresses the limitations of the current study and outlines directions for future research, particularly the need to account for the growing presence of autonomous driving technologies in collision analysis.
+The rest of the paper is organized as follows.
+Section 2 outlines the data and methodology.
+Section 3 presents the analysis of factors contributing to traffic collisions.
+Section 4 discusses the results and potential interventions.
+Section 5 addresses the study’s limitations and proposes future research directions, particularly the need to consider the growing role of autonomous driving technologies in accident analysis.
 
@@
 
-This traffic accident data consists of 18,957 entries from a citywide accident survey, capturing detailed information about various factors involved in traffic incidents.
-The raw dataset includes 50 variables, ranging from geographic data, such as the street where the accident occurred, to behavioral data, such as driver condition and alcohol involvement.
-This dataset records accidents over multiple years and provides insights into environmental conditions, road types, and the actions of drivers and pedestrians at the time of the incidents.
-The goal of this dataset is to provide a comprehensive view of traffic accidents for research and policy analysis related to urban traffic safety and accident prevention.
+This traffic accident dataset contains 18,957 entries from a citywide survey, detailing various factors involved in incidents.
+It includes 50 variables covering geographic data, such as accident locations, and behavioral data, like driver conditions and alcohol involvement.
+The dataset spans several years and captures environmental conditions, road types, and the actions of drivers and pedestrians during accidents.
+It aims to provide a comprehensive view for research and policy efforts related to urban traffic safety and accident prevention.
 
-The cleaned data is divided into two datasets, and the first one focuses on human factors related to traffic collisions, specifically the behaviors and conditions of drivers that may have contributed to accidents.
-The four main categories—'speeding', 'ag_driv', 'alcohol', and 'disability'—were selected because these are well-known, critical factors that significantly increase the risk of traffic accidents.
-Each factor is evaluated as either "None" or "Yes", indicating the absence or presence of that factor in the collision.
-These categories were chosen because they represent preventable human actions or conditions that are highly correlated with the likelihood and severity of traffic incidents.
-(see @tbl-cleaned-human-factors-data)
+The cleaned data is split into two sets.
+The first focuses on human factors, particularly driver behaviors and conditions that may have contributed to accidents.
+Four key categories—'speeding,' 'ag_driv,' 'alcohol,' and 'disability'—were selected due to their known impact on traffic accidents.
+Each is recorded as either "None" or "Yes," indicating whether that factor was present in the incident.
+These categories focus on preventable human actions or conditions that are strongly linked to traffic risks.
 
-1.
-**speeding**: Refers to collisions where speeding was a contributing factor.
-Speeding reduces a driver's ability to react quickly to road conditions or obstacles, and it increases the severity of crashes when they occur, making it one of the primary causes of traffic collisions.
+1. **Speeding**: Tracks whether speeding was a factor in the collision. Speeding reduces reaction time and increases crash severity, making it a leading cause of accidents.
 
-2.
-**ag_driv**: This category examines whether aggressive behavior, such as reckless driving, or distracted driving (like texting or talking on the phone) was involved in the collision.
-Aggressive and distracted driving are known to impair judgment and reaction time, leading to accidents that could otherwise have been avoided.
+2. **Aggressive Driving (ag_driv)**: Refers to reckless or distracted behaviors, such as texting while driving. These actions impair judgment and reaction time, significantly increasing accident risk.
 
-3.
-**alcohol**: Evaluates if alcohol was a factor in the collision.
-Alcohol impairs motor skills and decision-making, and is a well-documented contributor to severe and fatal traffic accidents.
-This category is crucial for understanding the role of substance use in driver behavior.
+3. **Alcohol**: Captures whether alcohol was involved in the collision. Alcohol impairs motor skills and decision-making, contributing to many severe and fatal accidents.
 
 4.
-**disability**: Assesses if any medical or physical disability played a role in the collision.
-Disabilities, whether temporary or permanent, can affect a driver's ability to control the vehicle or respond to changing road conditions, making this an important factor to consider in the context of traffic safety.
+**Disability**: Notes if a medical or physical disability affected the driver's ability to control the vehicle.
+Disabilities can impair a driver's capacity to respond to road conditions, making this a key safety factor.
 
-These categories were selected because they represent key human factors that have a direct impact on traffic safety, and addressing them can help reduce both the frequency and severity of accidents.
+These factors are critical to understanding the human behaviors that increase traffic accident risks and can inform strategies to reduce both the frequency and severity of collisions.
 
@@
 ```
-
-The second dataset focuses on the environmental factors influencing traffic collisions, capturing the conditions that shape driver behavior and contribute to accidents.
-These factors, including traffic control, visibility, lighting, and road surface conditions, play a vital role in either increasing or reducing the risk of collisions.
-The following modifications were made to simplify and group these categories, making the data more consistent and easier to analyze.
-The merged categories provide clearer insights into how these factors affect accidents and highlight areas for safety improvements.
+The second dataset focuses on environmental factors that influence traffic collisions, including conditions such as traffic control, visibility, lighting, and road surface.
+These elements shape driver behavior and impact accident risk.
+To simplify analysis and improve consistency, categories were consolidated, providing clearer insights into how these factors affect accidents and where safety improvements are needed.
 (see @tbl-cleaned-environmental-factors-data)
@@
 1.
-**traffic_control_type**: This category was simplified to include "No Control," "Traffic Signal," and "Others." The "Others" category consolidates less frequent types of traffic control, such as "Stop Sign," "Pedestrian Crossover," and "School Guard." This grouping reflects the most significant distinctions in how traffic is regulated, emphasizing scenarios where the presence or absence of formal traffic control could impact driver behavior.
-By focusing on the broad types of control, this category helps assess how different levels of traffic management influence accident risks.
+**Traffic Control Type**: This was simplified into "No Control," "Traffic Signal," and "Others." The "Others" group combines less common traffic controls like stop signs and pedestrian crossovers.
+This distinction highlights the impact of varying levels of traffic regulation on driver behavior, focusing on scenarios where traffic control presence may influence accident risks.
 
 2.
-**visibility**: The visibility conditions were grouped into "Clear" and "Disturbed." The "Disturbed" category includes various adverse weather conditions such as "Fog, Mist, Smoke, Dust," "Rain," "Snow," and "Strong Wind." This grouping simplifies the analysis by highlighting the key distinction between ideal and hazardous visibility conditions, which directly affect a driver’s ability to see and react to road conditions.
-Poor visibility is a known contributor to accidents, making this distinction critical for identifying environmental hazards.
+**Visibility**: Conditions were grouped into "Clear" and "Disturbed," with "Disturbed" covering adverse weather such as fog, rain, and snow.
+This division highlights the contrast between ideal and hazardous visibility, which directly affects a driver's ability to navigate safely.
+Poor visibility is a known factor in accidents, making this an essential category for identifying environmental hazards.
 
 3.
-**light**: The lighting conditions were merged into three categories: "Dark" (which includes "Dark" and "Dark, artificial"), "Daylight" (including "Daylight" and "Daylight, artificial"), and "Dim Light" (covering "Dawn," "Dusk," and similar intermediate lighting conditions).
-These groups reflect the significant lighting scenarios that impact driver visibility.
-Dark and dim lighting can reduce reaction times and perception, making these categories essential for understanding how lighting conditions contribute to accidents.
+**Light**: Lighting conditions were combined into "Dark," "Daylight," and "Dim Light." These reflect the most relevant lighting environments, with "Dark" including both natural and artificial lighting, and "Dim Light" covering intermediate conditions like dawn and dusk.
+Lighting plays a crucial role in visibility and reaction times, and this grouping helps assess its role in accidents.
 
 4.
-**road_surface_condition**: Road surface conditions were consolidated into "Dry," "Wet," and "Others." The "Others" category includes conditions like "Ice," "Loose Snow," and "Slush," which can severely reduce traction.
-Grouping less common conditions into "Others" helps streamline the analysis by focusing on the primary road conditions that affect vehicle control.
-Wet or slippery surfaces are known to increase accident risks, making this distinction important for understanding how road conditions influence collisions.
+**Road Surface Condition**: This was grouped into "Dry," "Wet," and "Others." The "Others" category includes ice, snow, and slush, conditions that greatly reduce vehicle traction.
+Focusing on major surface conditions helps clarify how road conditions influence accident risk, especially in wet or slippery environments.
 
-These categories were selected and modified to prioritize the most critical environmental factors that affect traffic safety.
-By simplifying and grouping related variables, the data becomes more interpretable, allowing for a clearer focus on how environmental factors like traffic control, visibility, lighting, and road conditions contribute to motor vehicle collisions.
-This approach not only aids in identifying preventable causes but also supports efforts to improve road safety and reduce collision rates through targeted interventions.
+These modifications prioritize the key environmental factors affecting traffic safety.
+By simplifying and grouping variables, the dataset becomes easier to interpret, allowing a clearer understanding of how conditions like traffic control, visibility, lighting, and road surfaces contribute to collisions.
+This approach supports targeted interventions aimed at improving road safety and reducing accident rates.
 
@@
 
-Lastly, the **acclass** category classifies the type and severity of traffic collisions with labels such as "Fatal," "Non-Fatal Injury," and "None or Others." The original classifications of "None" and "Property Damage Only" were merged into the "None or Others" category to streamline the data, as both involve accidents without human injury.
-This consolidation was taken to simplify the analysis by focusing on the human impact of accidents.
-"Fatal" represents collisions resulting in severe injury, and "Non-Fatal Injury" indicates incidents causing injury but no fatalities.
-By merging and refining these categories, the focus shifts toward understanding the human consequences of collisions, enabling a more targeted approach to reducing severe injuries and fatalities.
+The **acclass** category classifies the severity of traffic collisions into "Fatal," "Non-Fatal Injury," and "None or Others." The original categories "None" and "Property Damage Only" were combined into "None or Others" since both involve accidents without human injury.
+This consolidation simplifies the analysis, focusing on accidents with a direct human impact.
+"Fatal" refers to collisions causing severe injury or death, while "Non-Fatal Injury" includes incidents with injuries but no fatalities.
+By merging and refining these categories, the data emphasizes the human consequences of accidents, helping target efforts to reduce severe injuries and fatalities.
 
@@
 
-Upon further review of the heatmap, the top three combinations of environmental and human factors are evident, and I will revise the analysis accordingly.
+The heatmap shows the relationship between environmental factors—traffic control, visibility, lighting, and road surface—and human behaviors like speeding, aggressive driving, alcohol use, and disability.
+It highlights clear patterns where certain combinations lead to higher accident rates.
 
-### High-Risk Scenarios for Road Accidents
+The most dangerous combination occurs when **traffic signals are present, daylight is clear, the road is dry, and aggressive driving is involved**.
+Even in ideal environmental conditions, aggressive driving leads to the highest number of accidents.
+This suggests that behaviors like tailgating and reckless overtaking pose significant risks, even on dry roads with good visibility.
+To reduce accidents in such areas, stricter enforcement of speed limits and penalties for aggressive driving may be necessary.
 
-The heatmap highlights the intricate relationship between environmental conditions—such as traffic control, visibility, lighting, and road surface—and human factors like speeding, aggressive driving, alcohol use, and disability.
-This visualization reveals clear trends where specific combinations of these factors lead to the highest rates of road accidents.
+Another high-risk combination is **no traffic control, clear conditions, daylight, dry roads, and aggressive driving coupled with alcohol use**.
+The lack of traffic control in this scenario appears to give drivers a false sense of safety, leading to reckless behavior.
+Installing traffic controls, like stop signs or roundabouts, in areas with high accident rates could help reduce these incidents.
 
-The highest-risk combination occurs when **traffic signals are clear, daylight is present, the road is dry, and drivers exhibit aggressive driving behavior**.
-This combination results in the largest number of accidents.
-Even though environmental conditions are favorable with clear daylight and dry roads, aggressive driving significantly increases the accident rate.
-This indicates that aggressive behaviors, such as tailgating or reckless overtaking, are extremely dangerous even in otherwise safe driving conditions.
-To mitigate these risks, more targeted enforcement of speed limits and penalties for aggressive driving may be necessary in areas where accidents are high, despite good road conditions.
+A third combination showing increased risk involves **no traffic control, poor visibility, daylight, and wet roads**, but **no involvement of risky behaviors**.
+This highlights how wet roads and low visibility increase accident risks, even without human error.
+Improving road infrastructure, such as adding drainage systems or warning signs in areas prone to heavy rainfall, and introducing traffic controls could help prevent accidents under these conditions.
 
-A second high-risk combination involves **no traffic control, clear conditions, daylight, and dry roads, combined with aggressive driving and alcohol use**.
-This combination results in high accident counts, emphasizing how dangerous aggressive driving becomes when compounded by alcohol consumption, even in seemingly safe environmental conditions.
-The absence of traffic control in these situations suggests that drivers may feel a false sense of security, which encourages reckless behavior.
-Enhancing traffic control measures, such as adding stop signs, traffic lights, or roundabouts in high-accident zones, could reduce incidents in these cases.
-
-The third-highest combination includes **no traffic control, disturbed visibility, daylight, and wet road conditions** with **no involvement of impaired behaviors**.
-This scenario shows how environmental factors like wet roads and poor visibility increase the risk of accidents, even when human behavior is not a contributing factor.
-The lack of traffic control adds further risk.
-Addressing this issue through better traffic control measures and improving road surface conditions, such as better drainage systems or warning signs in areas prone to heavy rainfall, can help reduce accidents.
-
-The data reveals that while environmental factors such as wet roads, disturbed visibility, and lack of traffic control are significant contributors to accidents, aggressive driving and alcohol use are key human behaviors that exacerbate risks, even in favorable environmental conditions.
-Focusing on infrastructure improvements, such as enhancing traffic control in key areas and stricter enforcement against aggressive driving and alcohol-related offenses, could significantly reduce accident rates in the most hazardous scenarios.
+The data shows that while environmental factors like wet roads and poor visibility are important, aggressive driving and alcohol use are key behaviors that amplify risk, even in favorable conditions.
+Focusing on improving infrastructure and stricter enforcement against aggressive driving and alcohol-related offenses could significantly lower accident rates in these high-risk scenarios.
 
@@
 ## Can Environmental and Human Factors Serve as Early Warning Indicators?
-With the advancement of data-driven technologies, it is possible to predict the severity of motor vehicle accidents based on a combination of human behaviors and environmental conditions.
-This discussion focuses on how predictive models can be developed to identify the likelihood of severe accidents using factors like speeding, alcohol consumption, poor road conditions, and low visibility.
 
-For example, we could create a system that collects real-time environmental data (e.g., weather, road surface conditions) and cross-references it with information on traffic behaviors (e.g., average speeding rates) to predict whether an area is at high risk for serious accidents.
-This kind of early warning system could be integrated into navigation apps or smart cars to alert drivers of heightened risks and potentially prevent accidents before they occur.
+With advances in data-driven technologies, it's now possible to predict the severity of motor vehicle accidents by analyzing a combination of human behaviors and environmental conditions.
+This section focuses on how predictive models can be developed to assess the likelihood of severe accidents using factors such as speeding, alcohol use, poor road conditions, and low visibility.
 
-Predictive analytics can highlight the most dangerous combinations of factors, such as speeding in rainy weather or aggressive driving under poor visibility.
-In addition, machine learning algorithms can continuously improve these predictions by incorporating accident history data, making the model more robust over time.
-This offers the potential to shift from reactive to proactive accident prevention, reducing the burden on healthcare systems and increasing road safety.
+For instance, a system could gather real-time environmental data, such as weather and road surface conditions, and combine it with information on traffic behaviors, like speeding patterns, to determine if an area is at high risk for serious accidents.
+This type of early warning system could be integrated into navigation apps or smart cars to alert drivers to potential dangers and help prevent accidents before they occur.
+
+Predictive analytics can identify the most dangerous combinations of factors, such as speeding during rainy weather or aggressive driving in poor visibility.
+Machine learning algorithms could continuously improve these models by incorporating historical accident data, making the predictions more accurate over time.
+This approach shifts accident prevention from a reactive to a proactive strategy, which could reduce the strain on healthcare systems and improve overall road safety.
 
@@
 
-One notable limitation of this analysis is that the dataset only contains information up until 2023.
-While this provides a substantial historical perspective on traffic accidents in Toronto, it may not fully capture recent changes in road usage patterns, traffic enforcement, or driver behaviors that have emerged post-2023.
-For example, the rise of new driving technologies, ride-sharing services, and changes in traffic regulations could have a significant impact on accident rates, but these factors are not reflected in the current dataset.
-As such, the findings based on this data may not completely align with more recent trends, limiting the relevance of the analysis for current or future policy decisions.
+One limitation of this analysis is that the dataset only includes data up to 2023.
+While this offers a strong historical perspective on traffic accidents in Toronto, it does not account for more recent shifts in road usage, traffic enforcement, or driver behavior that may have emerged after 2023.
+For example, new driving technologies, ride-sharing services, and updated traffic regulations could significantly influence accident rates, but these factors are not reflected in the current data.
+As a result, the findings may not fully align with recent trends, limiting their relevance for current and future policy decisions.
 
-While the dataset used in this analysis ends in 2023, a key area for future exploration involves the broader adoption of autonomous or auto-drive technologies, which are not yet accounted for in the data but could significantly impact traffic accident dynamics.
-As these technologies become more prevalent, understanding their effect on traffic safety, accident rates, and accident severity will be critical.
-The absence of this data limits our ability to predict how autonomous driving might influence future accident patterns.
-However, from the figure "fig-trend-over-time," we observe that collisions involving aggressive driving (ag_driv) have decreased over time, which may be indicative of increased use of autonomous or driver-assist technologies that mitigate reckless behavior.
-Future research should incorporate data on autonomous vehicles and their interaction with human-driven cars to better assess how these advancements influence safety and accident risks.
-This could lead to the development of predictive models that combine human and automated driving behaviors, aiding policymakers and urban planners in shaping infrastructure and safety regulations for a future where both types of vehicles coexist.
+A key area for future research involves the growing use of autonomous or driver-assist technologies, which are not yet covered by the dataset but could have a major impact on traffic accident patterns.
+As these technologies become more common, understanding their effect on traffic safety and accident severity will be essential.
+The lack of data on autonomous vehicles limits our ability to predict how these innovations might change accident trends.
+However, from the figure @fig-trend-over-time, we see that accidents involving aggressive driving have decreased, possibly due to the rise of driver-assist features that reduce reckless behavior.
+Future studies should incorporate data on autonomous vehicles and their interactions with human-driven cars to better understand how these advancements affect safety.
+This could help develop predictive models that account for both human and automated driving behaviors, providing valuable insights for policymakers and urban planners as they adapt infrastructure and safety regulations for a future where both types of vehicles share the road.
 

### 2024-09-27 scripts/00-simulate_data.R
Update 00-simulate_data.R

@@ -62,7 +62,7 @@ set.seed(123)  # Setting seed for reproducibility
 
 # Simulate random data for traffic control type
 traffic_control <- tibble(
-  "Control Type" = c("No Control", "Traffic Signal", "Stop Sign", "Pedestrian Crossover"),
+  "Control Type" = c("No Control", "Traffic Signal", "Others"),
   "Count" = runif(4, 20, 60)  # Random counts between 20 and 60
 )
 
@@ -78,8 +78,8 @@ print(plot3)
 
 # Simulate random data for visibility conditions
 visibility <- tibble(
-  "Condition" = c("Clear", "Snow", "Rain"),
-  "Count" = runif(3, 30, 80)  # Random counts between 30 and 80
+  "Condition" = c("Clear", "Disturbed"),
+  "Count" = runif(2, 30, 80)  # Random counts between 30 and 80
 )
 
 # Plot: Visibility conditions
@@ -94,7 +94,7 @@ print(plot4)
 
 # Simulate random data for road surface conditions
 road_surface <- tibble(
-  "Condition" = c("Dry", "Ice", "Wet"),
+  "Condition" = c("Dry", "Others", "Wet"),
   "Count" = runif(3, 30, 80)  # Random counts between 30 and 80
 )
 
@@ -107,23 +107,4 @@ plot5 <- ggplot(road_surface, aes(x = Condition, y = Count)) +
 
 # Display road surface condition plot
 print(plot5)
-
-#### Simulation 4) Injury Results ####
-set.seed(123)  # Setting seed for reproducibility
-
-# Simulate random injury results
-injury_results <- tibble(
-  "Injury Type" = c("Fatal", "Major", "Minimal", "Minor", "None"),
-  "Count" = runif(5, 500, 9000)  # Random counts between 500 and 9000
-)
-
-# Plot: Injury results
-plot6 <- ggplot(injury_results, aes(x = `Injury Type`, y = Count, fill = `Injury Type`)) +
-  geom_bar(stat = "identity") +
-  labs(x = "Injury Type", y = "Count", title = "Injury Results") +
-  theme_minimal() +
-  geom_text(aes(label = round(Count)), vjust = -0.5)
-
-# Display injury results plot
-print(plot6)
 

### 2024-09-27 scripts/00-simulate_data.R
Update 00-simulate_data.R

@@ -11,37 +11,13 @@ library(tidyverse)
 library(ggplot2)
 library(tibble)
 
-#### Simulation 1) Human factor vs Environmental factor ####
-set.seed(123)  # Setting seed for reproducibility
-
-# Simulate random percentages for human vs environmental factors
-simulated_data1 <- tibble(
-  "Category" = c("Human", "Environmental"),
-  "Percentage" = runif(2, 20, 80)  # Generate random values between 20 and 80
-)
-
-# Normalize percentages to sum to 100
-simulated_data1 <- simulated_data1 %>%
-  mutate(Percentage = round(100 * Percentage / sum(Percentage)))
-
-# Plot: Pie chart to represent human vs environmental factors
-plot1 <- ggplot(simulated_data1, aes(x = "", y = Percentage, fill = Category)) +
-  geom_bar(stat = "identity", width = 1) +
-  coord_polar(theta = "y") +
-  labs(title = "Human vs Environmental Factors") +
-  theme_void() +
-  geom_text(aes(label = paste0(Percentage, "%")), position = position_stack(vjust = 0.5))
-
-# Display the pie chart
-print(plot1)
-
-#### Simulation 2) Human factors: AG-DRIV, Alcohol, Speeding ####
+#### Simulation 1) Human factors: AG-DRIV, Alcohol, Speeding ####
 set.seed(123)  # Setting seed for reproducibility
 
 # Simulate random data for human factors
 human_factors <- tibble(
-  "Factor" = c("AG-DRIV", "Alcohol", "Speeding"),
-  "Yes" = runif(3, 30, 80),  # Random 'Yes' values between 30 and 80
+  "Factor" = c("AG-DRIV", "Alcohol", "Speeding", "Disability"),
+  "Yes" = runif(4, 30, 80),  # Random 'Yes' values between 30 and 80
   "No" = 100 - Yes  # 'No' is complementary to 'Yes' so they sum to 100
 )
 
@@ -49,32 +25,32 @@ human_factors <- tibble(
 human_factors_long <- human_factors %>%
   pivot_longer(cols = c("Yes", "No"), names_to = "Response", values_to = "Count")
 
-plot2 <- ggplot(human_factors_long, aes(x = Factor, y = Count, fill = Response)) +
+plot1 <- ggplot(human_factors_long, aes(x = Factor, y = Count, fill = Response)) +
   geom_bar(stat = "identity", position = "dodge") +
[…]

### 2024-09-27 paper/paper.qmd
final revision

@@
 title: "Analysis of Motor Vehicle Collision Causes in Toronto from 2006 to 2023"
-subtitle: "Exploring the Interplay of Human Behavior and Environmental Factors in Predicting Accident Severity and High-Risk Scenarios"
+subtitle: "Assessing the Interplay of Human Behavior and Environmental Factors in Predicting Accident Severity and High-Risk Scenarios"
 author: 
@@
 library(tidyverse)
+library(ggplot2)
+library(dplyr)
+library(readr)
 library(knitr)
 library(here)
-library(dplyr)
-library(ggplot2)
-library(readr)
 library(gridExtra)
@@
 Motor vehicle collisions are a leading cause of injury and fatality in urban areas, driving ongoing efforts to improve road safety.
-In a rapidly expanding city like Toronto, understanding the causes of traffic accidents is critical for shaping effective policies.
-This paper examines traffic collisions in Toronto over a 17-year period, focusing on how human behavior and environmental conditions together influence the frequency and severity of accidents.
+In an expanding city like Toronto, understanding the causes of traffic accidents is critical for shaping effective policies.
+This paper examines traffic collisions in Toronto over 17 years, focusing on how human behavior and environmental conditions influence the frequency and severity of accidents.
 
@@
 Most studies focus on these factors separately, missing how they combine to increase accident risks.
-This research addresses that gap by analyzing both human and environmental factors simultaneously, offering a more complete view of what drives traffic accidents in an urban setting.
+This research addresses that gap by analyzing both human and environmental factors simultaneously, offering a more complete view of what leads to traffic accidents in an urban setting.
 
@@
 High-risk scenarios, like aggressive driving on dry roads with clear visibility, emerge as leading causes of collisions.
-These findings suggest targeted actions, such as stricter enforcement against reckless driving and improved traffic controls in key areas, could help reduce accident rates.
+These findings suggest that targeted actions, such as stricter enforcement against reckless driving and improved traffic controls in key areas, could help reduce accident rates.
 
 The rest of the paper is organized as follows.
-Section 2 outlines the data and methodology.
-Section 3 presents the analysis of factors contributing to traffic collisions.
-Section 4 discusses the results and potential interventions.
-Section 5 addresses the study’s limitations and proposes future research directions, particularly the need to consider the growing role of autonomous driving technologies in accident analysis.
+@sec-data outlines the data and methodology and analyzes factors contributing to traffic collisions.
+@sec-discussion discusses the results and potential interventions.
+The final sections of the discussion part critically evaluate the study's limitations and propose directions for future research.
 
@@
 
-## Data Source
+## Data Source and Collection
 
 This research is based on a dataset published by Toronto Police Services, which is accessible through the City of Toronto’s OpenDataToronto Library [@rOpenDataToronto].
-The specific dataset includes all traffic collisions events in Toronto where a person was either Killed or Seriously Injured (KSI) since 2006.
-
-## Data Collection
-
-The methodology for gathering data is structured to ensure a detailed and current perspective of shelter and overnight service.
-The data was compiled and examined using the R statistical programming software [@r], supplemented by various tools such as `tidyverse` [@rTidyverse], `ggplot2` [@rGgplot2], `dplyr` [@rDplyr], `readr` [@rReadr], `tibble` [@rTibble], `janitor` [@rJanitor], `knitr` [@rknitr], and `here` [@rHere].
+The specific dataset includes all traffic collision events in Toronto where a person was either Killed or Seriously Injured (KSI) since 2006.
+The data was compiled and examined using the R statistical programming software [@r], supplemented by various tools such as `tidyverse` [@rTidyverse], `ggplot2` [@rGgplot2], `dplyr` [@rDplyr], `readr` [@rReadr], `gridExtra` [@gridExtra], `grid` [@R-base-grid], `knitr` [@rknitr], and `here` [@rHere].
 
@@
 The dataset spans several years and captures environmental conditions, road types, and the actions of drivers and pedestrians during accidents.
-It aims to provide a comprehensive view for research and policy efforts related to urban traffic safety and accident prevention.
+It aims to provide a comprehensive view of research and policy efforts related to urban traffic safety and accident prevention.
+Although the dataset was last updated September 27, 2024, but only contains observations from 2006 to 2023.
+
+The `acclass` category classifies the severity of traffic collisions into "Fatal," "Non-Fatal Injury," and "None or Others." The original categories "None" and "Property Damage Only" were combined into "None or Others" since both involve accidents without human injury.
+This consolidation simplifies the analysis, focusing on accidents with a direct human impact.
+"Fatal" refers to collisions causing severe injury or death, while "Non-Fatal Injury" includes incidents with injuries but no fatalities.
+The accident classification data shows that most collisions result in non-fatal injuries (16,268), while fatal accidents account for 2,670 cases.
+The 'None or Others' category has the fewest incidents, with only 19 cases recorded.
+This highlights the high frequency of non-fatal injuries, emphasizing the need for policies focused on injury prevention.
+(see @fig-acclass-distribution)
+
+```{r}
+#| message: false
+#| echo: false
+#| label: fig-acclass-distribution
+#| fig-cap: Distribution of Accident Classifications
+
+# Ensure the 'acclass' column is treated as a factor
+cleaned_data <- cleaned_data %>%
+  mutate(acclass = factor(acclass, levels = c("Fatal", "Non-Fatal Injury", "None or Others")))
+
+# Summarise the counts for each 'acclass' category
+acclass_summary <- cleaned_data %>%
+  count(acclass)
+
+# Create the bar plot
+ggplot(acclass_summary, aes(x = acclass, y = n, fill = acclass)) +
+  geom_bar(stat = "identity") +
+  geom_text(aes(label = n), vjust = -0.5, nudge_y = 200, size = 3, fontface = "plain", color = "black") +
+  theme_minimal() +
+  labs(title = "Distribution of Accident Classifications",
+       x = "Accident Classification (acclass)",
+       y = "Count") +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
+  coord_cartesian(ylim = c(0, max(acclass_summary$n) * 1.1))
+
+```
+
+## Human Factors
 
 The cleaned data is split into two sets.
-The first focuses on human factors, particularly driver behaviors and conditions that may have contributed to accidents.
-Four key categories—'speeding,' 'ag_driv,' 'alcohol,' and 'disability'—were selected due to their known impact on traffic accidents.
+The first focuses on human factors, particularly driver behaviors, and conditions that may have contributed to accidents.
+Four key categories—`speeding,` `ag_driv,` `alcohol` and `disability`—were selected due to their known impact on traffic accidents.
 Each is recorded as either "None" or "Yes," indicating whether that factor was present in the incident.
 These categories focus on preventable human actions or conditions that are strongly linked to traffic risks.
+(see @tbl-cleaned-human-factors-data)
+
+```{r}
+#| message: false
+#| echo: false
+#| label: tbl-cleaned-human-factors-data
+#| tbl-cap: Sample of Human Factors of Collision Data
+
+head (environmental_factors_data, 5) |>
+  kable(
+    col.names = c("date", "speeding", "ag_driv", "alcohol", "disability"),
+  booktabs = TRUE
+  )
+```
 
@@
 
-These factors are critical to understanding the human behaviors that increase traffic accident risks and can inform strategies to reduce both the frequency and severity of collisions.
+@fig-human-factors-frequency-distribution illustrates the frequency of human factors contributing to traffic collisions.
+Aggressive and disturbed driving (ag_driv) appears most frequently, with 9,836 occurrences, followed by speeding, which accounts for 2,694 cases.
+Alcohol use and disability are less common, with 808 and 493 incidents, respectively.
+Understanding these factors is important for identifying the human behaviors that increase traffic accident risks and for shaping strategies to reduce both the frequency and severity of collisions.
 
@@
 #| echo: false
-#| label: fig-frequency-factors-tall
-#| fig-cap: Frequency of human factors with thinner numbers and taller graph
+#| label: fig-human-factors-frequency-distribution
+#| fig-cap: Frequency of Human Factors in Traffic Collisions
 
@@
 
-# Split the category name for 'ag_driv'
-factor_summary <- factor_summary %>%
-  mutate(factor = str_replace_all(factor, "ag_driv", "aggressive\nand disturbed driving"))
-
 # Reorder factors based on count for better visualization
@@
 
-```{r}
-#| message: false
-#| echo: false
-#| label: fig-trend-over-time
-#| fig-cap: Trends of human factors in traffic collisions over time
+## Environmental Factors
 
-# Convert the date column to date format
-human_factors_data <- human_factors_data %>%
-  mutate(date = as.Date(date, format = "%Y-%m-%d"))
-
-# Summarise the presence of each factor by year
-trend_data <- human_factors_data %>%
-  mutate(year = lubridate::year(date)) %>%
-  group_by(year) %>%
-  summarise(
-    speeding = sum(speeding == "Yes"),
-    ag_driv = sum(ag_driv == "Yes"),
-    alcohol = sum(alcohol == "Yes"),
-    disability = sum(disability == "Yes")
-  ) %>%
-  tidyr::gather(key = "factor", value = "count", -year)
-
-# Line plot to show trends over time
-ggplot(trend_data, aes(x = year, y = count, color = factor)) +
-  geom_line(linewidth = 0.8) +
-  theme_minimal() +
-  labs(title = "Trends in Human Factors Over Time",
-       x = "Year",
-       y = "Count of Collisions",
-       color = "Human Factor")
-```
 The second dataset focuses on environmental factors that influence traffic collisions, including conditions such as traffic control, visibility, lighting, and road surface.
 These elements shape driver behavior and impact accident risk.
-To simplify analysis and improve consistency, categories were consolidated, providing clearer insights into how these factors affect accidents and where safety improvements are needed.
+To simplify analysis and improve consistency, categories were consolidated, providing a clearer understanding of how these factors affect accidents and where safety improvements are needed.
 (see @tbl-cleaned-environmental-factors-data)
-
-1.
-**Traffic Control Type**: This was simplified into "No Control," "Traffic Signal," and "Others." The "Others" group combines less common traffic controls like stop signs and pedestrian crossovers.
-This distinction highlights the impact of varying levels of traffic regulation on driver behavior, focusing on scenarios where traffic control presence may influence accident risks.
-
-2.
-**Visibility**: Conditions were grouped into "Clear" and "Disturbed," with "Disturbed" covering adverse weather such as fog, rain, and snow.
-This division highlights the contrast between ideal and hazardous visibility, which directly affects a driver's ability to navigate safely.
-Poor visibility is a known factor in accidents, making this an essential category for identifying environmental hazards.
-
-3.
-**Light**: Lighting conditions were combined into "Dark," "Daylight," and "Dim Light." These reflect the most relevant lighting environments, with "Dark" including both natural and artificial lighting, and "Dim Light" covering intermediate conditions like dawn and dusk.
-Lighting plays a crucial role in visibility and reaction times, and this grouping helps assess its role in accidents.
-
-4.
-**Road Surface Condition**: This was grouped into "Dry," "Wet," and "Others." The "Others" category includes ice, snow, and slush, conditions that greatly reduce vehicle traction.
-Focusing on major surface conditions helps clarify how road conditions influence accident risk, especially in wet or slippery environments.
-
-These modifications prioritize the key environmental factors affecting traffic safety.
-By simplifying and grouping variables, the dataset becomes easier to interpret, allowing a clearer understanding of how conditions like traffic control, visibility, lighting, and road surfaces contribute to collisions.
-This approach supports targeted interventions aimed at improving road safety and reducing accident rates.
 
@@
 
+1.
+**Traffic Control Type (traffctl)**: This was simplified into "No Control," "Traffic Signal," and "Others." The "Others" group combines less common traffic controls like stop signs and pedestrian crossovers.
+This distinction highlights the impact of varying levels of traffic regulation on driver behavior, focusing on scenarios where traffic control presence may influence accident risks.
+
+2.
+**Visibility**: Conditions were grouped into "Clear" and "Disturbed," with "Disturbed" covering adverse weather such as fog, rain, and snow.
+This division highlights the contrast between ideal and hazardous visibility, which directly affects a driver's ability to navigate safely.
+Poor visibility is a known factor in accidents, making this an essential category for identifying environmental hazards.
+
+3.
+**Light**: Lighting conditions were combined into "Dark," "Daylight," and "Dim Light." These reflect the most relevant lighting environments, with "Dark" including both natural and artificial lighting, and "Dim Light" covering intermediate conditions like dawn and dusk.
+Lighting affects visibility and reaction times, and this grouping helps assess its role in accidents.
+
+4.
+**Road Surface Condition (rdsfcond)**: This was grouped into "Dry," "Wet," and "Others." The "Others" category includes ice, snow, and slush, conditions that greatly reduce vehicle traction.
+Focusing on major surface conditions helps clarify how road conditions influence accident risk, especially in wet or slippery environments.
+
+@fig-environmental-factors-traffic-control-visibility and @fig-environmental-factors-light-road-surface show that there were a similar number of collisions with no traffic control, totaling 9,021, and with traffic signals, at 8,035.
+The majority occurred under clear visibility with 16,373 collisions, during daylight with 10,930 collisions, and on dry road surfaces with 15,231 collisions.
+In contrast, there were 2,584 collisions under disturbed visibility and 729 under dim light conditions.
+Wet surfaces accounted for 3,140 collisions, while dark conditions saw 7,298 collisions, highlighting the importance of environmental and visibility factors in traffic accidents.
+
 ```{r}
@@
 #| echo: false
-#| label: fig-four-graphs-layout
-#| fig-cap: Combined layout with four simplified collision graphs
+#| label: fig-environmental-factors-traffic-control-visibility
+#| fig-cap: Environmental Factors Affecting Collisions - Traffic Control and Visibility
 
@@
 
-# Plot 2: Visibility Conditions
+# Plot 2: Visibility Condition
 plot2 <- ggplot(environmental_factors_data, aes(x = visibility)) +
@@
   common_theme +
-  labs(title = "Collisions by Visibility Conditions",
+  labs(title = "Collisions by Visibility",
        x = "Visibility",
@@
 first_plot_pair <- grid.arrange(plot1, plot2, ncol = 2, 
-                                top = textGrob("Environmental Factors Affecting Collisions", gp = gpar(fontsize = 15, fontface = "bold")))
+                                top = textGrob("Environmental Factors Affecting Collisions: Traffic Control and Visibility", gp = gpar(fontsize = 10, fontface = "bold")))
 
@@
 #| echo: false
-#| label: tbl-cleaned-environmental-factors-data-2
-#| tbl-cap: Sample of Environmental Factors of Collision Data 2
+#| label: fig-environmental-factors-light-road-surface
+#| fig-cap: Environmental Factors Affecting Collisions - Light and Road Surface
 
@@
   common_theme +  # Apply the common theme
-  labs(title = "Collisions by Lighting Conditions",
-       x = "Lighting Conditions",
+  labs(title = "Collisions by light",
+       x = "Lighting Condition",
        y = "Collisions") +
@@
   common_theme +  # Apply the common theme
-  labs(title = "Collisions by Road Surface Condition",
+  labs(title = "Collisions by rdsfcond",
        x = "Road Surface Condition",
@@
 second_plot_pair <- grid.arrange(plot3, plot4, ncol = 2, 
-                                 top = textGrob("Environmental Factors Affecting Collisions: Part 2", gp = gpar(fontsize = 15, fontface = "bold")))
+                                 top = textGrob("Environmental Factors Affecting Collisions: Light and Road Surface Condition", gp = gpar(fontsize = 10, fontface = "bold")))
 
@@
 
-The **acclass** category classifies the severity of traffic collisions into "Fatal," "Non-Fatal Injury," and "None or Others." The original categories "None" and "Property Damage Only" were combined into "None or Others" since both involve accidents without human injury.
-This consolidation simplifies the analysis, focusing on accidents with a direct human impact.
-"Fatal" refers to collisions causing severe injury or death, while "Non-Fatal Injury" includes incidents with injuries but no fatalities.
-By merging and refining these categories, the data emphasizes the human consequences of accidents, helping target efforts to reduce severe injuries and fatalities.
-
-```{r}
-#| message: false
-#| echo: false
-#| label: fig-acclass-distribution
-#| fig-cap: Distribution of Accident Classifications (acclass)
-
-# Ensure the 'acclass' column is treated as a factor
-cleaned_data <- cleaned_data %>%
-  mutate(acclass = factor(acclass, levels = c("Fatal", "Non-Fatal Injury", "None or Others")))
-
-# Summarise the counts for each 'acclass' category
-acclass_summary <- cleaned_data %>%
-  count(acclass)
-
-# Create the bar plot
-ggplot(acclass_summary, aes(x = acclass, y = n, fill = acclass)) +
-  geom_bar(stat = "identity") +
-  geom_text(aes(label = n), vjust = -0.5, nudge_y = 200, size = 3, fontface = "plain", color = "black") +
-  theme_minimal() +
-  labs(title = "Distribution of Accident Classifications",
-       x = "Accident Classification",
-       y = "Count") +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
-  coord_cartesian(ylim = c(0, max(acclass_summary$n) * 1.1))  # Adjust y-axis limits for better readability
-
-```
-
-# Discussion
+# Discussion {#sec-discussion}
 
@@
 
-The heatmap shows the relationship between environmental factors—traffic control, visibility, lighting, and road surface—and human behaviors like speeding, aggressive driving, alcohol use, and disability.
+@fig-high-risk-combinations-for-accident shows the relationship between environmental factors—traffic control, visibility, lighting, and road surface—and human behaviors like speeding, aggressive driving, alcohol use, and disability.
 It highlights clear patterns where certain combinations lead to higher accident rates.
@@
 #| echo: false
-#| label: fig-hh
-#| fig-cap: Sample of Result of Collision Data
+#| label: fig-high-risk-combinations-for-accident
+#| fig-cap: High-Risk Scenarios for Collisions
 # Prepare the data by summarizing relevant columns and creating a count of combinations
@@
 ## Can Environmental and Human Factors Serve as Early Warning Indicators?
+```{r}
+#| message: false
+#| echo: false
+#| label: fig-human-factor-trend-over-time
+#| fig-cap: Trends of human factors in traffic collisions over time
+
+# Convert the date column to date format
+human_factors_data <- human_factors_data %>%
+  mutate(date = as.Date(date, format = "%Y-%m-%d"))
+
+# Summarise the presence of each factor by year
+trend_data <- human_factors_data %>%
+  mutate(year = lubridate::year(date)) %>%
+  group_by(year) %>%
+  summarise(
+    speeding = sum(speeding == "Yes"),
+    ag_driv = sum(ag_driv == "Yes"),
+    alcohol = sum(alcohol == "Yes"),
+    disability = sum(disability == "Yes")
+  ) %>%
+  tidyr::gather(key = "factor", value = "count", -year)
+
+# Line plot to show trends over time
+ggplot(trend_data, aes(x = year, y = count, color = factor)) +
+  geom_line(linewidth = 0.8) +
+  theme_minimal() +
+  labs(title = "Trends in Human Factors Over Time",
+       x = "Year",
+       y = "Count of Collisions",
+       color = "Human Factor")
+```
 
@@
 One limitation of this analysis is that the dataset only includes data up to 2023.
-While this offers a strong historical perspective on traffic accidents in Toronto, it does not account for more recent shifts in road usage, traffic enforcement, or driver behavior that may have emerged after 2023.
+While this offers a strong historical view on traffic accidents in Toronto, it does not account for more recent shifts in road usage, traffic enforcement, or driver behavior that may have emerged after 2023.
 For example, new driving technologies, ride-sharing services, and updated traffic regulations could significantly influence accident rates, but these factors are not reflected in the current data.
@@
 The lack of data on autonomous vehicles limits our ability to predict how these innovations might change accident trends.
-However, from the figure @fig-trend-over-time, we see that accidents involving aggressive driving have decreased, possibly due to the rise of driver-assist features that reduce reckless behavior.
+However, from the figure @fig-human-factor-trend-over-time, we see that accidents involving aggressive driving have decreased, possibly due to the rise of driver-assist features that reduce reckless behavior.
 Future studies should incorporate data on autonomous vehicles and their interactions with human-driven cars to better understand how these advancements affect safety.
-This could help develop predictive models that account for both human and automated driving behaviors, providing valuable insights for policymakers and urban planners as they adapt infrastructure and safety regulations for a future where both types of vehicles share the road.
+This could help develop predictive models that account for both human and automated driving behaviors, providing important understanding for policymakers and urban planners as they adapt infrastructure and safety regulations for a future where both types of vehicles share the road.
 

### 2024-09-27 paper/paper.qmd
final final revision!!

@@
 title: "Analysis of Motor Vehicle Collision Causes in Toronto from 2006 to 2023"
-subtitle: "Assessing the Interplay of Human Behavior and Environmental Factors in Predicting Accident Severity and High-Risk Scenarios"
+subtitle: "Assessing the Interplay of Human behaviour and Environmental Factors in Predicting Accident Severity and High-Risk Scenarios"
 author: 
@@
 date-format: long
-abstract: "This study analyzes motor vehicle collisions in Toronto from 2006 to 2023, focusing on the interaction between human behaviors—such as speeding, aggressive driving, and alcohol use—and environmental conditions like visibility and road surfaces.
-The findings show that while environmental factors increase risks, aggressive driving and alcohol use are the strongest predictors of severe accidents.
-These results highlight the need for targeted interventions in both driver behavior and infrastructure to improve urban traffic safety."
+abstract: "This study analyzes motor vehicle collisions in Toronto from 2006 to 2023, focusing on the interaction between human behaviours—such as speeding, aggressive driving, and alcohol use—and environmental conditions like visibility and road surfaces.
+The findings show that while environmental factors increase risks, aggressive driving is the strongest predictors of severe accidents.
+These results highlight the need for targeted interventions in both driver behaviour and infrastructure to improve urban traffic safety."
 format: pdf
@@
 In an expanding city like Toronto, understanding the causes of traffic accidents is critical for shaping effective policies.
-This paper examines traffic collisions in Toronto over 17 years, focusing on how human behavior and environmental conditions influence the frequency and severity of accidents.
+This paper examines traffic collisions in Toronto over 17 years, focusing on how human behaviour and environmental conditions influence the frequency and severity of accidents.
 
-While much research has been done on factors such as speeding or poor road conditions, there is still a lack of understanding regarding the interaction between human behaviors and environmental conditions in creating high-risk situations.
+While much research has been done on factors such as speeding or poor road conditions, there is still a lack of understanding regarding the interaction between human behaviours and environmental conditions in creating high-risk situations.
 Most studies focus on these factors separately, missing how they combine to increase accident risks.
@@
 The cleaned data is split into two sets.
-The first focuses on human factors, particularly driver behaviors, and conditions that may have contributed to accidents.
+The first focuses on human factors, particularly driver behaviours, and conditions that may have contributed to accidents.
 Four key categories—`speeding,` `ag_driv,` `alcohol` and `disability`—were selected due to their known impact on traffic accidents.
@@
 
-2. **Aggressive Driving (ag_driv)**: Refers to reckless or distracted behaviors, such as texting while driving. These actions impair judgment and reaction time, significantly increasing accident risk.
+2.
+**Aggressive Driving (ag_driv)**: Refers to reckless or distracted behaviours, such as texting while driving.
+These actions impair judgment and reaction time, significantly increasing accident risk.
 
@@
 Alcohol use and disability are less common, with 808 and 493 incidents, respectively.
-Understanding these factors is important for identifying the human behaviors that increase traffic accident risks and for shaping strategies to reduce both the frequency and severity of collisions.
+Understanding these factors is important for identifying the human behaviours that increase traffic accident risks and for shaping strategies to reduce both the frequency and severity of collisions.
 
@@
 The second dataset focuses on environmental factors that influence traffic collisions, including conditions such as traffic control, visibility, lighting, and road surface.
-These elements shape driver behavior and impact accident risk.
+These elements shape driver behaviour and impact accident risk.
 To simplify analysis and improve consistency, categories were consolidated, providing a clearer understanding of how these factors affect accidents and where safety improvements are needed.
@@
 **Traffic Control Type (traffctl)**: This was simplified into "No Control," "Traffic Signal," and "Others." The "Others" group combines less common traffic controls like stop signs and pedestrian crossovers.
-This distinction highlights the impact of varying levels of traffic regulation on driver behavior, focusing on scenarios where traffic control presence may influence accident risks.
+This distinction highlights the impact of varying levels of traffic regulation on driver behaviour, focusing on scenarios where traffic control presence may influence accident risks.
 
@@
 
-@fig-high-risk-combinations-for-accident shows the relationship between environmental factors—traffic control, visibility, lighting, and road surface—and human behaviors like speeding, aggressive driving, alcohol use, and disability.
+@fig-high-risk-combinations-for-accident shows the relationship between environmental factors—traffic control, visibility, lighting, and road surface—and human behaviours like speeding, aggressive driving, alcohol use, and disability.
 It highlights clear patterns where certain combinations lead to higher accident rates.
 
-The most dangerous combination occurs when **traffic signals are present, daylight is clear, the road is dry, and aggressive driving is involved**.
+A fatal collision frequently occurs when traffic signals are present, clear visibility, daylight, the road is dry, and aggressive driving is involved.
 Even in ideal environmental conditions, aggressive driving leads to the highest number of accidents.
-This suggests that behaviors like tailgating and reckless overtaking pose significant risks, even on dry roads with good visibility.
+This suggests that behaviours like tailgating and reckless overtaking pose significant risks, even on dry roads with good visibility.
 To reduce accidents in such areas, stricter enforcement of speed limits and penalties for aggressive driving may be necessary.
 
-Another high-risk combination is **no traffic control, clear conditions, daylight, dry roads, and aggressive driving coupled with alcohol use**.
-The lack of traffic control in this scenario appears to give drivers a false sense of safety, leading to reckless behavior.
+Another high-risk combination is no traffic control, clear visibility, daylight, and dry roads with no human factors involved.
+The lack of traffic control in this scenario appears to give drivers a false sense of safety, leading to reckless behaviour.
 Installing traffic controls, like stop signs or roundabouts, in areas with high accident rates could help reduce these incidents.
 
-A third combination showing increased risk involves **no traffic control, poor visibility, daylight, and wet roads**, but **no involvement of risky behaviors**.
-This highlights how wet roads and low visibility increase accident risks, even without human error.
-Improving road infrastructure, such as adding drainage systems or warning signs in areas prone to heavy rainfall, and introducing traffic controls could help prevent accidents under these conditions.
-
-The data shows that while environmental factors like wet roads and poor visibility are important, aggressive driving and alcohol use are key behaviors that amplify risk, even in favorable conditions.
+This shows that while environmental factors like no traffic controls, wet roads and poor visibility are important, aggressive driving is a key behaviour that amplifies risk, even in favorable conditions.
 Focusing on improving infrastructure and stricter enforcement against aggressive driving and alcohol-related offenses could significantly lower accident rates in these high-risk scenarios.
@@
 
-With advances in data-driven technologies, it's now possible to predict the severity of motor vehicle accidents by analyzing a combination of human behaviors and environmental conditions.
+With advances in data-driven technologies, it's now possible to predict the severity of motor vehicle accidents by analyzing a combination of human behaviours and environmental conditions.
 This section focuses on how predictive models can be developed to assess the likelihood of severe accidents using factors such as speeding, alcohol use, poor road conditions, and low visibility.
 
-For instance, a system could gather real-time environmental data, such as weather and road surface conditions, and combine it with information on traffic behaviors, like speeding patterns, to determine if an area is at high risk for serious accidents.
+For instance, a system could gather real-time environmental data, such as weather and road surface conditions, and combine it with information on traffic behaviours, like speeding patterns, to determine if an area is at high risk for serious accidents.
 This type of early warning system could be integrated into navigation apps or smart cars to alert drivers to potential dangers and help prevent accidents before they occur.
@@
 
-One limitation of this analysis is that the dataset only includes data up to 2023.
-While this offers a strong historical view on traffic accidents in Toronto, it does not account for more recent shifts in road usage, traffic enforcement, or driver behavior that may have emerged after 2023.
+One limitation of this analysis is that the dataset only includes observations up to 2023.
+While this offers a strong historical view on traffic accidents in Toronto, it does not account for more recent shifts in road usage, traffic enforcement, or driver behaviour that may have emerged after 2023.
 For example, new driving technologies, ride-sharing services, and updated traffic regulations could significantly influence accident rates, but these factors are not reflected in the current data.
@@
 The lack of data on autonomous vehicles limits our ability to predict how these innovations might change accident trends.
-However, from the figure @fig-human-factor-trend-over-time, we see that accidents involving aggressive driving have decreased, possibly due to the rise of driver-assist features that reduce reckless behavior.
+However, from the figure @fig-human-factor-trend-over-time, we see that accidents involving aggressive driving have decreased, possibly due to the rise of driver-assist features that reduce reckless behaviour.
 Future studies should incorporate data on autonomous vehicles and their interactions with human-driven cars to better understand how these advancements affect safety.
-This could help develop predictive models that account for both human and automated driving behaviors, providing important understanding for policymakers and urban planners as they adapt infrastructure and safety regulations for a future where both types of vehicles share the road.
+This could help develop predictive models that account for both human and automated driving behaviours, providing important understanding for policymakers and urban planners as they adapt infrastructure and safety regulations for a future where both types of vehicles share the road.
 
